---
title: "milt — Modern Integrated Library for Timeseries"
subtitle: "Complete Interactive Documentation"
author: "Gideon Nti Boateng"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: flatly
    code-fold: false
    code-tools: true
    highlight-style: github
    embed-resources: true
    smooth-scroll: true
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  message: false
  cache: false
---

```{r setup}
#| include: false
# Load the package from source (works without installing)
devtools::load_all(quiet = TRUE)
```

# Introduction

**milt** (Modern Integrated Library for Timeseries) is a unified R framework for time series analysis. Think Python's *Darts* or *sktime*, but natively in R with a tidyverse-friendly API.

## What Can milt Do?

| Capability | Functions |
|------------------------------------|------------------------------------|
| **Create & manipulate series** | `milt_series()`, `milt_split()`, `milt_window()`, … |
| **Forecast** | `milt_model()` → `milt_fit()` → `milt_forecast()` |
| **Anomaly detection** | `milt_detector()` → `milt_detect()` |
| **Change-point analysis** | `milt_changepoints()` |
| **Classification** | `milt_classifier()` → `milt_classify_fit()` → `milt_classify_predict()` |
| **Clustering** | `milt_cluster()` |
| **Feature engineering** | `milt_pipeline()` + step DSL |
| **Explainability** | `milt_explain()` |
| **Causal impact** | `milt_causal_impact()` |
| **Deployment** | `milt_serve()`, `milt_dashboard()`, `milt_report()`, `milt_save()` |

## Core Design Principles

**Unified pipe-first API.** Every function takes its main object as the first argument so you can chain with `|>`:

``` r
milt_model("ets") |> milt_fit(series) |> milt_forecast(horizon = 12)
```

**S3 outside, R6 inside.** You never call `$methods()` directly. Use the exported verbs. The R6 objects live inside and manage state.

**Lazy dependencies.** Heavy packages (`torch`, `xgboost`, `prophet`, …) are only loaded when you use the backend that needs them — you get a friendly error message if the package is missing.

------------------------------------------------------------------------

# The MiltSeries — Your Data Container {#sec-series}

Everything in milt starts with a `MiltSeries` object. It wraps your data with metadata: a time column, a value column, an optional grouping column, and the series frequency.

## Creating a Series

### From a `ts` object (most common)

```{r series-from-ts}
# AirPassengers is a built-in R dataset — monthly airline passengers 1949-1960
ap <- milt_series(AirPassengers)
print(ap)
```

You can also pass the frequency and start explicitly:

```{r series-from-ts-explicit}
ap2 <- milt_series(AirPassengers, frequency = 12, start = c(1949, 1))
print(ap2)
```

### From a numeric vector

```{r series-from-vector}
x <- milt_series(
  rnorm(60, mean = 100, sd = 10),
  frequency = 12,
  start     = c(2018, 1)
)
print(x)
```

### From a data frame / tibble

```{r series-from-df}
df <- data.frame(
  date  = seq(as.Date("2020-01-01"), by = "month", length.out = 36),
  sales = cumsum(rnorm(36, 5, 2)) + 100
)

s_df <- milt_series(df, time_col = "date", value_col = "sales", frequency = 12)
print(s_df)
```

## Inspecting a Series

```{r series-inspect}
ap <- milt_series(AirPassengers)

# Basic metadata
ap$n_timesteps()    # number of observations
ap$freq()           # "monthly"
ap$start_time()     # first timestamp
ap$end_time()       # last timestamp
ap$values()         # raw numeric vector
```

```{r series-tibble}
# Convert to a tibble for dplyr workflows
tibble::as_tibble(ap) |> head()
```

## Subsetting and Slicing

```{r series-slice}
# Head / tail
milt_head(ap, 6)
milt_tail(ap, 6)
```

```{r series-window}
# Extract a date window
milt_window(ap,
  start = as.Date("1955-01-01"),
  end   = as.Date("1957-12-01")
)
```

```{r series-bracket}
# [ operator — same as milt_window() but by position
ap[1:24]
```

## Train / Test Split

```{r series-split}
# 80% train, 20% test
spl <- milt_split(ap, ratio = 0.8)
cat("Train:", spl$train$n_timesteps(), "obs\n")
cat("Test: ", spl$test$n_timesteps(),  "obs\n")
```

```{r series-split-at}
# Split at a specific date
spl2 <- milt_split_at(ap, time = as.Date("1958-01-01"))
cat("Train ends:", format(spl2$train$end_time()), "\n")
cat("Test starts:", format(spl2$test$start_time()), "\n")
```

## Concatenation and Resampling

```{r series-concat}
s1 <- ap[1:60]
s2 <- ap[61:144]
recombined <- milt_concat(s1, s2)
recombined$n_timesteps()
```

```{r series-resample}
# Resample to quarterly (from monthly)
ap_q <- milt_resample(ap, period = "quarterly", agg_fn = sum)
print(ap_q)
```

## Handling Missing Values

```{r series-fill-gaps}
# Introduce a gap then fill it
vals <- ap$values()
vals[c(10, 11, 12)] <- NA
s_na <- milt_series(vals, frequency = 12, start = c(1949, 1))

# Fill using linear interpolation
s_filled <- milt_fill_gaps(s_na, method = "linear")
print(s_filled)
```

## Visualising a Series

```{r series-plot}
plot(ap)
```

------------------------------------------------------------------------

# Exploratory Data Analysis {#sec-eda}

`milt_eda()` gives you a quick statistical profile of your series — trend direction, seasonality, stationarity, and outlier counts.

```{r eda}
ap <- milt_series(AirPassengers)
eda_result <- milt_eda(ap)
print(eda_result)
```

## ACF and Decomposition Plots

```{r plot-acf}
milt_plot_acf(ap, lag.max = 36L)
```

```{r plot-decomp}
milt_plot_decomp(ap)
```

## Series Diagnostics with Recommendations

`milt_diagnose()` runs a suite of statistical tests and prints model recommendations.

```{r diagnose}
milt_diagnose(ap)
```

------------------------------------------------------------------------

# Forecasting {#sec-forecasting}

## The Universal Forecasting Interface

milt uses the same three-step pipeline for every model:

```         
milt_model("<name>")  →  milt_fit(<series>)  →  milt_forecast(horizon = <h>)
```

Let's see what models are available:

```{r list-models}
list_milt_models()
```

## Classical Statistical Models

### ARIMA

Auto-ARIMA automatically selects the best (p, d, q) order.

```{r arima}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)

arima_fit <- milt_model("auto_arima") |> milt_fit(ap)
arima_fct <- milt_forecast(arima_fit, horizon = 24)

print(arima_fct)
plot(arima_fct)
```

### ETS (Exponential Smoothing)

```{r ets}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ets_fct <- milt_model("ets") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 24)

plot(ets_fct)
```

### STL Decomposition

```{r stl}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
stl_fct <- milt_model("stl") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 24)

plot(stl_fct)
```

### Theta Method

```{r theta}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
theta_fct <- milt_model("theta") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(theta_fct)
```

### TBATS

Good for complex seasonality (e.g. daily data with weekly + annual cycles).

```{r tbats}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
tbats_fct <- milt_model("tbats") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(tbats_fct)
```

### Prophet (Facebook)

```{r prophet}
#| eval: !expr requireNamespace("prophet", quietly = TRUE)
prophet_fct <- milt_model("prophet") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 24)

plot(prophet_fct)
```

### Croston (Intermittent Demand)

Croston's method is designed for intermittent (sparse) demand series.

```{r croston}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Simulate intermittent demand
demand <- c(0,0,3,0,0,0,2,0,5,0,0,1,0,0,0,4,0,0,2,0,
            0,0,1,0,3,0,0,0,2,0,0,1,0,0,5,0)
demand_series <- milt_series(demand, frequency = 12, start = c(2020, 1))

croston_fct <- milt_model("croston") |>
  milt_fit(demand_series) |>
  milt_forecast(horizon = 12)

print(croston_fct)
```

### Naive Baseline

Always a good baseline to beat.

```{r naive}
naive_fct <- milt_model("naive") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(naive_fct)
```

## Machine Learning Models

ML models automatically build lag features from the series to train a regression model. You can tune how many lags to use.

### XGBoost

```{r xgboost}
#| eval: !expr requireNamespace("xgboost", quietly = TRUE)
xgb_fct <- milt_model("xgboost", lags = 1:24) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

plot(xgb_fct)
```

### LightGBM

```{r lightgbm}
#| eval: !expr requireNamespace("lightgbm", quietly = TRUE)
lgb_fct <- milt_model("lightgbm", lags = 1:24) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(lgb_fct)
```

### Random Forest

```{r random-forest}
#| eval: !expr requireNamespace("ranger", quietly = TRUE)
rf_fct <- milt_model("random_forest", lags = 1:12) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

plot(rf_fct)
```

### Elastic Net

```{r elastic-net}
#| eval: !expr requireNamespace("glmnet", quietly = TRUE)
en_fct <- milt_model("elastic_net", lags = 1:12) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(en_fct)
```

### Support Vector Machine

```{r svm}
#| eval: !expr requireNamespace("e1071", quietly = TRUE)
svm_fct <- milt_model("svm", lags = 1:12) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(svm_fct)
```

### K-Nearest Neighbours

```{r knn}
knn_fct <- milt_model("knn", lags = 1:12) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(knn_fct)
```

## Deep Learning Models (torch)

Deep learning models require the `torch` package with its Lantern C++ backend installed.

``` r
# One-time setup — downloads the C++ runtime
install.packages("torch")
torch::install_torch()
```

All DL backends follow the same pattern:

``` r
#| eval: false
# N-BEATS
nbeats_fct <- milt_model("nbeats", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# N-HiTS
nhits_fct <- milt_model("nhits", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# Temporal Convolutional Network (TCN)
tcn_fct <- milt_model("tcn", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# Temporal Fusion Transformer (TFT)
tft_fct <- milt_model("tft", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# PatchTST
patch_fct <- milt_model("patch_tst", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)
```

### Checking the torch device

``` r
milt_torch_device()  # returns cuda if GPU available, else cpu
```

------------------------------------------------------------------------

# Working with Forecasts {#sec-forecast-objects}

A `MiltForecast` object stores the full forecast result — point estimates, prediction intervals, and (for probabilistic models) sample paths.

```{r forecast-object}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)
fct <- milt_model("ets") |> milt_fit(ap) |> milt_forecast(horizon = 24)

# Inspect
print(fct)
plot(fct)
```

```{r forecast-to-tibble}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Extract as a tibble for further processing
tibble::as_tibble(fct) |> head()
```

## In-sample Predictions and Residuals

```{r predict-residuals}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
m <- milt_model("ets") |> milt_fit(ap)

fitted_vals <- milt_predict(m)
residuals   <- milt_residuals(m)

plot(residuals, type = "l", main = "Residuals", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```

------------------------------------------------------------------------

# Accuracy Metrics {#sec-metrics}

milt ships a comprehensive suite of forecast accuracy functions.

```{r metrics-setup}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap   <- milt_series(AirPassengers)
spl  <- milt_split(ap, ratio = 0.8)
fct  <- milt_model("ets") |> milt_fit(spl$train) |> milt_forecast(spl$test$n_timesteps())

actual <- spl$test$values()
pred   <- tibble::as_tibble(fct)$.mean
```

```{r metrics-all}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Comprehensive accuracy report
milt_accuracy(actual, pred)
```

```{r metrics-individual}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Individual metrics
cat("MAE:  ", milt_mae(actual, pred), "\n")
cat("RMSE: ", milt_rmse(actual, pred), "\n")
cat("MAPE: ", milt_mape(actual, pred), "%\n")
cat("sMAPE:", milt_smape(actual, pred), "%\n")
cat("MASE: ", milt_mase(actual, pred, spl$train$values()), "\n")
cat("R²:   ", milt_r_squared(actual, pred), "\n")
```

## Probabilistic Metrics

```{r probabilistic-metrics}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# For probabilistic forecasts — need sample paths
# (Using simulated samples here for illustration)
samples <- matrix(rnorm(length(actual) * 200, mean = pred, sd = 15),
                  nrow = length(actual), ncol = 200)

cat("CRPS:    ", milt_crps(actual, samples), "\n")
cat("Coverage:", milt_coverage(actual, pred - 1.96*15, pred + 1.96*15), "\n")
```

------------------------------------------------------------------------

# Backtesting and Model Evaluation {#sec-backtest}

## Simple Backtest

```{r backtest}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)

bt <- milt_backtest(
  milt_model("ets"),
  series         = ap,
  initial_window = 84,
  horizon        = 12,
  stride         = 12L
)

print(bt)
tibble::as_tibble(bt) |> head()
```

## Cross-Validation

```{r crossval}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
cv_result <- milt_cv(
  milt_model("auto_arima"),
  series         = ap,
  initial_window = 72,
  horizon        = 12
)

print(cv_result)
```

## Comparing Multiple Models

```{r compare}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
cmp <- milt_compare(
  models = list(
    arima = milt_model("auto_arima"),
    ets   = milt_model("ets"),
    naive = milt_model("naive")
  ),
  series  = ap,
  horizon = 12
)

print(cmp)
plot(cmp)
```

------------------------------------------------------------------------

# Ensemble Models {#sec-ensemble}

Combine multiple models with averaging or weighted combinations.

```{r ensemble}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)

m1 <- milt_model("auto_arima") |> milt_fit(ap)
m2 <- milt_model("ets")        |> milt_fit(ap)
m3 <- milt_model("naive")      |> milt_fit(ap)

ens <- milt_ensemble(
  list(arima = m1, ets = m2, naive = m3),
  weights = c(0.5, 0.4, 0.1)
)

ens_fct <- milt_forecast(ens, horizon = 12)
plot(ens_fct)
```

------------------------------------------------------------------------

# Feature Engineering with Pipelines {#sec-pipelines}

The pipeline system lets you compose data transformations into a reusable, fittable object. Pipelines fit on training data and can be applied to new data — preventing data leakage.

## Building a Pipeline

A pipeline always needs a model attached as its final step. Use `milt_pipe_model()` to attach one.

```{r pipeline-build}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)

pipe <- milt_pipeline() |>
  milt_pipe_step_lag(lags = 1:12) |>
  milt_pipe_step_rolling(windows = c(3, 6, 12), fns = c("mean", "sd")) |>
  milt_pipe_step_fourier(K = 3) |>
  milt_pipe_step_calendar() |>
  milt_pipe_step_scale() |>
  milt_pipe_model(milt_model("naive"))

print(pipe)
```

## Fitting and Transforming

```{r pipeline-fit}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
spl         <- milt_split(ap, ratio = 0.8)
fitted_pipe <- milt_pipeline_fit(pipe, spl$train)

# Forecast using the fitted pipeline
pipe_fct <- milt_pipeline_forecast(fitted_pipe, horizon = 12)
print(pipe_fct)
```

## Pipeline + Model

Attach a model as the final step of the pipeline:

```{r pipe-model}
#| eval: !expr requireNamespace("xgboost", quietly = TRUE)
full_pipe <- milt_pipeline() |>
  milt_pipe_step_lag(lags = 1:12) |>
  milt_pipe_step_rolling(windows = c(3, 6), fns = "mean") |>
  milt_pipe_step_scale() |>
  milt_pipe_model(milt_model("xgboost"))

pipe_fit <- milt_pipeline_fit(full_pipe, spl$train)
pipe_fct <- milt_pipeline_forecast(pipe_fit, horizon = 12)
plot(pipe_fct)
```

------------------------------------------------------------------------

# Individual Feature Steps {#sec-steps}

The `milt_step_*` functions work directly on a `MiltSeries` without a full pipeline.

```{r step-scale}
ap <- milt_series(AirPassengers)

# Scale the series (z-score normalisation)
scaled <- milt_step_scale(ap)
print(scaled)
```

```{r step-lag}
# Add lag features (returns tibble)
lag_df <- milt_step_lag(ap, lags = c(1, 3, 12))
head(lag_df)
```

```{r step-rolling}
# Rolling statistics
roll_df <- milt_step_rolling(ap, windows = 6, fns = "mean")
head(roll_df)
```

```{r step-fourier}
# Fourier terms for seasonality
fourier_df <- milt_step_fourier(ap, K = 2)
head(fourier_df)
```

```{r step-calendar}
# Calendar features (month, quarter, weekday, …)
cal_df <- milt_step_calendar(ap)
head(cal_df)
```

------------------------------------------------------------------------

# Anomaly Detection {#sec-anomaly}

## The Detector Interface

Like models, detectors follow a two-step interface:

```         
milt_detector("<method>")  →  milt_detect(<series>)
```

## Statistical Methods

```{r detect-iqr}
ap <- milt_series(AirPassengers)

# IQR-based detection (classical box-plot rule)
iqr_result <- milt_detector("iqr") |> milt_detect(ap)
print(iqr_result)
plot(iqr_result)
```

```{r detect-stl}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# STL decomposition — flags residuals outside ±3 SD
stl_result <- milt_detector("stl") |> milt_detect(ap)
print(stl_result)
```

```{r detect-grubbs}
#| eval: !expr requireNamespace("tseries", quietly = TRUE)
# Grubbs test (univariate extreme value)
grubbs_result <- milt_detector("grubbs") |> milt_detect(ap)
print(grubbs_result)
```

```{r detect-gesd}
# Generalised ESD test (multiple outliers)
gesd_result <- milt_detector("gesd") |> milt_detect(ap)
print(gesd_result)
```

## Machine Learning Methods

```{r detect-lof}
#| eval: !expr requireNamespace("dbscan", quietly = TRUE)
# Local Outlier Factor
lof_result <- milt_detector("lof") |> milt_detect(ap)
print(lof_result)
```

```{r detect-iforest}
#| eval: !expr requireNamespace("isotree", quietly = TRUE)
# Isolation Forest
iforest_result <- milt_detector("isolation_forest") |> milt_detect(ap)
print(iforest_result)
```

```{r detect-autoencoder}
#| eval: !expr requireNamespace("torch", quietly = TRUE)
# Autoencoder (deep learning)
ae_result <- milt_detector("autoencoder") |> milt_detect(ap)
print(ae_result)
```

## Ensemble Detection

Combine multiple detectors — a point is anomalous only if a majority vote says so.

```{r detect-ensemble}
ens_result <- milt_detector("ensemble",
  detectors = list(
    milt_detector("iqr"),
    milt_detector("gesd")
  )
) |> milt_detect(ap)
print(ens_result)
```

## MiltAnomalies Result Object

```{r anomalies-object}
result <- milt_detector("iqr") |> milt_detect(ap)

# Convert to tibble
tibble::as_tibble(result) |> dplyr::filter(is_anomaly) |> head()
```

------------------------------------------------------------------------

# Change-Point Detection {#sec-changepoints}

```{r changepoints}
#| eval: !expr requireNamespace("changepoint", quietly = TRUE)
ap <- milt_series(AirPassengers)

cp_result <- milt_changepoints(ap, method = "pelt")
print(cp_result)
plot(cp_result)
```

------------------------------------------------------------------------

# Multi-Series Analysis {#sec-multi-series}

## Creating a Multi-Series Object

```{r multi-series-create}
# Simulate three store sales series
set.seed(42)
sales_df <- data.frame(
  date  = rep(seq(as.Date("2019-01-01"), by = "month", length.out = 48), 3),
  store = rep(c("A", "B", "C"), each = 48),
  sales = c(
    100 + cumsum(rnorm(48, 1, 3)),
    200 + cumsum(rnorm(48, 2, 5)),
    150 + cumsum(rnorm(48, 0.5, 2))
  )
)

ms <- milt_series(sales_df,
  time_col  = "date",
  value_col = "sales",
  group_col = "store",
  frequency = 12
)

print(ms)
```

## Fitting Models Across All Groups

```{r multi-fit}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Fit ETS to every store independently
ms_fit <- milt_model("ets") |> milt_fit(ms)
ms_fct <- milt_forecast(ms_fit, horizon = 12)
print(ms_fct)
```

------------------------------------------------------------------------

# Hierarchical Reconciliation {#sec-hierarchical}

When your series form a hierarchy (e.g., country → region → store), forecasts at lower levels should sum to the forecast at the higher level. `milt_reconcile()` enforces this.

```{r hierarchical}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
set.seed(42)
# Three individual series: total, region_A, region_B
make_hier <- function(n, mean_drift) {
  milt_series(300 + cumsum(rnorm(n, mean_drift, 3)),
              frequency = 12, start = c(2019, 1))
}

s_total    <- make_hier(48, 3)
s_region_a <- make_hier(48, 1)
s_region_b <- make_hier(48, 2)

# Fit ETS independently on each series
fct_total    <- milt_model("ets") |> milt_fit(s_total)    |> milt_forecast(horizon = 12)
fct_region_a <- milt_model("ets") |> milt_fit(s_region_a) |> milt_forecast(horizon = 12)
fct_region_b <- milt_model("ets") |> milt_fit(s_region_b) |> milt_forecast(horizon = 12)

# Summation matrix S: rows = [total, region_A, region_B], cols = [region_A, region_B]
S <- matrix(
  c(1, 1,   # total  = region_A + region_B
    1, 0,   # region_A
    0, 1),  # region_B
  nrow = 3, ncol = 2, byrow = TRUE
)

# Reconcile — forecasts is a named list of MiltForecast objects
reconciled <- milt_reconcile(
  forecasts = list(total = fct_total, region_A = fct_region_a, region_B = fct_region_b),
  S         = S,
  method    = "ols"
)
print(reconciled)
tibble::as_tibble(reconciled)
```

------------------------------------------------------------------------

# Time Series Classification {#sec-classify}

Classify time series into predefined categories. Useful for product type detection, fault classification, customer segmentation, and more.

```{r classify}
#| eval: !expr requireNamespace("ranger", quietly = TRUE)
# Build labelled training data
set.seed(1)
make_series <- function(n, trend, noise) {
  milt_series(cumsum(rnorm(n, trend, noise)),
    frequency = 1, start = c(2020, 1))
}

# Three classes: trending up, flat, trending down
train_series <- list(
  s1 = make_series(50,  0.5, 1),
  s2 = make_series(50,  0.0, 1),
  s3 = make_series(50, -0.5, 1),
  s4 = make_series(50,  0.6, 1),
  s5 = make_series(50, -0.4, 1)
)

labels <- c("up", "flat", "down", "up", "down")

# Fit classifier
clf <- milt_classifier("random_forest") |>
  milt_classify_fit(train_series, labels)

# Predict on new series
new_series <- list(
  t1 = make_series(50,  0.4, 1),
  t2 = make_series(50, -0.3, 1)
)

predictions <- milt_classify_predict(clf, new_series)
print(predictions)
```

------------------------------------------------------------------------

# Time Series Clustering {#sec-cluster}

Group similar time series together without labels.

```{r cluster}
#| eval: !expr requireNamespace("dtw", quietly = TRUE)
# Simulate 9 series from 3 underlying patterns
set.seed(42)
series_list <- lapply(1:9, function(i) {
  group   <- (i - 1) %% 3
  trend   <- c(0.3, 0, -0.3)[group + 1]
  vals    <- cumsum(rnorm(60, trend, 1))
  milt_series(vals, frequency = 12, start = c(2018, 1))
})

cl_result <- milt_cluster(series_list, k = 3, method = "dtw")
print(cl_result)
```

------------------------------------------------------------------------

# Explainability {#sec-explain}

`milt_explain()` uses SHAP-style feature importance to show which lag/feature values drove each forecast.

```{r explain}
#| eval: !expr requireNamespace("xgboost", quietly = TRUE)
ap <- milt_series(AirPassengers)

xgb_model <- milt_model("xgboost", lags = 1:12) |> milt_fit(ap)
explanation <- milt_explain(xgb_model, ap)

print(explanation)
plot(explanation)
```

------------------------------------------------------------------------

# Causal Impact Analysis {#sec-causal}

Measure the effect of an intervention (e.g., a marketing campaign, product launch, policy change) by building a counterfactual forecast of what *would have* happened without it.

```{r causal-impact}
#| eval: !expr requireNamespace("CausalImpact", quietly = TRUE)
set.seed(1)
n        <- 100
pre_n    <- 70
baseline <- cumsum(rnorm(n, 0.5, 1)) + 100

# Add a treatment effect from observation 71 onwards
treatment                       <- baseline
treatment[(pre_n + 1):n] <- treatment[(pre_n + 1):n] + 15

dates  <- seq(as.Date("2016-01-01"), by = "week", length.out = n)
series <- milt_series(treatment, frequency = 52, start = c(2016, 1))

ci_result <- milt_causal_impact(
  series,
  event_time = dates[pre_n + 1]
)

print(ci_result)
plot(ci_result)
```

------------------------------------------------------------------------

# Saving and Loading Models {#sec-save-load}

Trained models can be serialised to disk and reloaded in a fresh session.

```{r save-load}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap    <- milt_series(AirPassengers)
model <- milt_model("ets") |> milt_fit(ap)

# Save to a file
tmp <- tempfile(fileext = ".milt")
milt_save(model, path = tmp)

# Load from file
loaded <- milt_load(tmp)
fct    <- milt_forecast(loaded, horizon = 12)
print(fct)

# Clean up
file.remove(tmp)
```

------------------------------------------------------------------------

# Model Registry {#sec-registry}

milt has a global model registry. You can inspect it and register your own custom model backends.

```{r registry-list}
# Inspect all registered models
models_tbl <- list_milt_models()
print(models_tbl)
```

## Registering a Custom Model

You can register any R6 class that follows the `MiltModelBase` interface:

``` r
# Minimal custom model skeleton
MyCustomModel <- R6::R6Class(
  classname = "MyCustomModel",
  inherit   = MiltModelBase,
  public = list(
    fit = function(series) {
      # ... fit logic ...
      private$.is_fitted <- TRUE
      invisible(self)
    },
    forecast = function(horizon, ...) {
      # ... forecast logic ...
    },
    predict = function(series = NULL, ...) {
      # ... in-sample prediction ...
    },
    residuals = function(...) {
      # ... residuals ...
    }
  )
)

# Register it
register_milt_model(
  name        = "my_model",
  class       = MyCustomModel,
  description = "My custom model"
)

# Use it just like any built-in model
milt_model("my_model") |> milt_fit(ap) |> milt_forecast(horizon = 12)
```

------------------------------------------------------------------------

# Deployment {#sec-deployment}

## Generating a Report

`milt_report()` renders an HTML / PDF report from a fitted forecast.

``` r
# Requires pandoc and rmarkdown
ap    <- milt_series(AirPassengers)
model <- milt_model("ets") |> milt_fit(ap)

milt_report(
  series      = ap,
  models      = list(ets = model),
  horizon     = 24L,
  output_format = "html",
  output_file   = "my_forecast_report.html"
)
```

## Interactive Dashboard

`milt_dashboard()` launches a Shiny dashboard to explore forecasts interactively.

``` r
# Requires shiny
milt_dashboard(
  model  = model,
  series = ap
)
```

## REST API with Plumber

`milt_serve()` wraps a trained model as a REST API using the `plumber` package.

``` r
# Requires plumber
api <- milt_serve(model, port = 8080)

# Then call your model:
# POST http://localhost:8080/predict  {"horizon": 12}
```

------------------------------------------------------------------------

# Deep Dive: The Architecture {#sec-architecture}

Understanding how milt is structured helps when extending it.

## 4-Layer Architecture

```         
┌──────────────────────────────────────────────────┐
│  Layer 1: User API                               │
│  milt_model(), milt_fit(), milt_forecast(), …    │
│  S3 generics: print(), plot(), summary()         │
├──────────────────────────────────────────────────┤
│  Layer 2: Core Engine                            │
│  MiltSeriesR6, MiltForecastR6, ModelRegistry     │
│  MiltAnomaliesR6, MiltBacktestR6, …              │
├──────────────────────────────────────────────────┤
│  Layer 3: Model Backends                         │
│  backend-arima.R, backend-ets.R, backend-xgb.R  │
│  backend-nbeats.R, detect-iqr.R, …              │
├──────────────────────────────────────────────────┤
│  Layer 4: Data Layer                             │
│  tsibble, arrow, data.table, tibble              │
└──────────────────────────────────────────────────┘
```

## Adding a New Model Backend

1.  Create `R/backend-mymodel.R`
2.  Define an R6 class inheriting `MiltModelBase`
3.  Implement `fit()`, `forecast()`, `predict()`, `residuals()`
4.  Call `register_milt_model()` at package load time
5.  Done — no changes needed anywhere else

## The Dependency Strategy

| Type | Packages | Behaviour |
|------------------------|------------------------|------------------------|
| **Imports** | `R6`, `rlang`, `cli`, `dplyr`, `tibble`, `tsibble`, … | Always available |
| **Suggests** | `forecast`, `xgboost`, `torch`, `prophet`, … | Loaded on demand via `check_installed_backend()` |

------------------------------------------------------------------------

# Complete End-to-End Example {#sec-e2e}

Let's walk through a complete forecasting workflow — from raw data to deployed forecast — in one coherent example.

```{r e2e-full}
#| eval: !expr (requireNamespace("forecast", quietly = TRUE) && requireNamespace("xgboost", quietly = TRUE))

# ── 1. Create the series ──────────────────────────────────────────────────────
ap <- milt_series(AirPassengers, frequency = 12, start = c(1949, 1))
print(ap)

# ── 2. Explore ────────────────────────────────────────────────────────────────
eda_result <- milt_eda(ap)
print(eda_result)

# ── 3. Diagnose ───────────────────────────────────────────────────────────────
milt_diagnose(ap)

# ── 4. Split ──────────────────────────────────────────────────────────────────
spl <- milt_split(ap, ratio = 0.8)

# ── 5. Detect anomalies on training data ─────────────────────────────────────
anomalies <- milt_detector("iqr") |> milt_detect(spl$train)
print(anomalies)

# ── 6. Fit and compare three models ──────────────────────────────────────────
cmp <- milt_compare(
  models = list(
    ets   = milt_model("ets"),
    arima = milt_model("auto_arima"),
    knn   = milt_model("knn", lags = 1:12)
  ),
  series  = spl$train,
  horizon = spl$test$n_timesteps()
)
print(cmp)

# ── 7. Pick best and forecast ─────────────────────────────────────────────────
best_model <- milt_model("ets") |> milt_fit(ap)
final_fct  <- milt_forecast(best_model, horizon = 24)

plot(final_fct)

# ── 8. Score on held-out test set ─────────────────────────────────────────────
test_fct <- milt_model("ets") |>
  milt_fit(spl$train) |>
  milt_forecast(spl$test$n_timesteps())

milt_accuracy(spl$test$values(), tibble::as_tibble(test_fct)$.mean)
```

------------------------------------------------------------------------

# Quick Reference {#sec-reference}

## Series Functions

| Function                      | Purpose                     |
|-------------------------------|-----------------------------|
| `milt_series()`               | Create a MiltSeries         |
| `milt_head()` / `milt_tail()` | First / last N observations |
| `milt_window()`               | Subset by date range        |
| `milt_split()`                | Train/test split by ratio   |
| `milt_split_at()`             | Train/test split by date    |
| `milt_concat()`               | Join two series             |
| `milt_resample()`             | Change frequency            |
| `milt_fill_gaps()`            | Interpolate missing values  |
| `milt_to_tibble()`            | Convert to tibble           |
| `milt_to_ts()`                | Convert to base R `ts`      |
| `milt_to_tsibble()`           | Convert to tsibble          |

## Model Functions

| Function                        | Purpose                       |
|---------------------------------|-------------------------------|
| `list_milt_models()`            | Show all registered models    |
| `milt_model(name)`              | Instantiate a model           |
| `milt_fit(model, series)`       | Train the model               |
| `milt_forecast(model, horizon)` | Generate forecasts            |
| `milt_predict(model)`           | In-sample fitted values       |
| `milt_residuals(model)`         | Training residuals            |
| `milt_backtest()`               | Rolling-window backtest       |
| `milt_cv()`                     | Time-series cross-validation  |
| `milt_compare()`                | Side-by-side model comparison |
| `milt_ensemble()`               | Weighted model ensemble       |

## Model Names

| Name              | Type          | Package  |
|-------------------|---------------|----------|
| `"auto_arima"`    | Classical     | forecast |
| `"ets"`           | Classical     | forecast |
| `"stl"`           | Classical     | forecast |
| `"theta"`         | Classical     | forecast |
| `"tbats"`         | Classical     | forecast |
| `"prophet"`       | Classical     | prophet  |
| `"croston"`       | Classical     | forecast |
| `"naive"`         | Baseline      | —        |
| `"xgboost"`       | ML            | xgboost  |
| `"lightgbm"`      | ML            | lightgbm |
| `"random_forest"` | ML            | ranger   |
| `"elastic_net"`   | ML            | glmnet   |
| `"svm"`           | ML            | e1071    |
| `"knn"`           | ML            | —        |
| `"nbeats"`        | Deep Learning | torch    |
| `"nhits"`         | Deep Learning | torch    |
| `"tcn"`           | Deep Learning | torch    |
| `"tft"`           | Deep Learning | torch    |
| `"patch_tst"`     | Deep Learning | torch    |
| `"deepar"`        | Deep Learning | torch    |

## Anomaly Detectors

| Name                 | Method                      |
|----------------------|-----------------------------|
| `"iqr"`              | IQR box-plot rule           |
| `"stl"`              | STL decomposition residuals |
| `"grubbs"`           | Grubbs extreme-value test   |
| `"gesd"`             | Generalised ESD test        |
| `"lof"`              | Local Outlier Factor        |
| `"isolation_forest"` | Isolation Forest            |
| `"autoencoder"`      | Deep learning autoencoder   |
| `"ensemble"`         | Majority-vote combination   |

## Metrics

| Function           | Description                         |
|--------------------|-------------------------------------|
| `milt_accuracy()`  | All metrics at once                 |
| `milt_mae()`       | Mean Absolute Error                 |
| `milt_mse()`       | Mean Squared Error                  |
| `milt_rmse()`      | Root Mean Squared Error             |
| `milt_mape()`      | Mean Absolute % Error               |
| `milt_smape()`     | Symmetric MAPE                      |
| `milt_mase()`      | Mean Abs Scaled Error               |
| `milt_rmsse()`     | Root Mean Sq Scaled Error           |
| `milt_r_squared()` | R²                                  |
| `milt_crps()`      | Continuous Ranked Probability Score |
| `milt_coverage()`  | Prediction interval coverage        |
| `milt_pinball()`   | Pinball / quantile loss             |
| `milt_winkler()`   | Winkler interval score              |

------------------------------------------------------------------------

# Session Info {#sec-session}

```{r session-info}
sessionInfo()
```
