---
title: "milt — Modern Integrated Library for Timeseries"
subtitle: "Complete Interactive Documentation"
author: "Gideon Nti Boateng"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: flatly
    code-fold: false
    code-tools: true
    highlight-style: github
    embed-resources: true
    smooth-scroll: true
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  message: false
  cache: false
---

```{r setup}
#| include: false
library(milt)
```

# Introduction

**milt** (Modern Integrated Library for Timeseries) is a unified R framework for time series analysis. Think Python's *Darts* or *sktime*, but natively in R with a tidyverse-friendly API.

## What Can milt Do?

| Capability | Functions |
|----|----|
| **Create & manipulate series** | `milt_series()`, `milt_split()`, `milt_window()`, … |
| **TimeSeries operations** | arithmetic (`+`/`-`/`*`/`/`), `milt_stack()`, `milt_step_map()`, `milt_add_datetime_component()`, `milt_add_holidays()` |
| **Forecast** | `milt_model()` → `milt_fit()` → `milt_forecast()` |
| **Filter (smooth observed data)** | `milt_filter()` — moving-average, Kalman, Gaussian Process |
| **Backtest & search hyperparameters** | `milt_backtest()`, `milt_cv()`, `milt_grid_search()` |
| **Calibrated intervals for any model** | `milt_conformal()` |
| **Anomaly detection** | `milt_detector()` → `milt_detect()` |
| **Change-point analysis** | `milt_changepoints()` |
| **Classification** | `milt_classifier()` → `milt_classify_fit()` → `milt_classify_predict()` |
| **Clustering** | `milt_cluster()` |
| **Feature engineering** | `milt_pipeline()` + step DSL (lags, rolling, Fourier, calendar, scale, Box-Cox, diff, map) |
| **Explainability** | `milt_explain()` |
| **Causal impact** | `milt_causal_impact()` |
| **Deployment** | `milt_serve()`, `milt_dashboard()`, `milt_report()`, `milt_save()` |

## Core Design Principles

**Unified pipe-first API.** Every function takes its main object as the first argument so you can chain with `|>`:

``` r
milt_model("ets") |> milt_fit(series) |> milt_forecast(horizon = 12)
```

**S3 outside, R6 inside.** You never call `$methods()` directly. Use the exported verbs. The R6 objects live inside and manage state.

**Lazy dependencies.** Heavy packages (`torch`, `xgboost`, `prophet`, …) are only loaded when you use the backend that needs them — you get a friendly error message if the package is missing.

------------------------------------------------------------------------

# The MiltSeries — Your Data Container {#sec-series}

Everything in milt starts with a `MiltSeries` object. It wraps your data with metadata: a time column, a value column, an optional grouping column, and the series frequency.

## Creating a Series

### From a `ts` object (most common)

```{r series-from-ts}
# AirPassengers is a built-in R dataset — monthly airline passengers 1949-1960
ap <- milt_series(AirPassengers)
print(ap)
```

You can also pass the frequency and start explicitly:

```{r series-from-ts-explicit}
ap2 <- milt_series(AirPassengers, frequency = 12, start = c(1949, 1))
print(ap2)
```

### From a numeric vector

```{r series-from-vector}
x <- milt_series(
  rnorm(60, mean = 100, sd = 10),
  frequency = 12,
  start     = c(2018, 1)
)
print(x)
```

### From a data frame / tibble

```{r series-from-df}
df <- data.frame(
  date  = seq(as.Date("2020-01-01"), by = "month", length.out = 36),
  sales = cumsum(rnorm(36, 5, 2)) + 100
)

s_df <- milt_series(df, time_col = "date", value_col = "sales", frequency = 12)
print(s_df)
```

## Inspecting a Series

```{r series-inspect}
ap <- milt_series(AirPassengers)

# Basic metadata
ap$n_timesteps()    # number of observations
ap$freq()           # "monthly"
ap$start_time()     # first timestamp
ap$end_time()       # last timestamp
ap$values()         # raw numeric vector
```

```{r series-tibble}
# Convert to a tibble for dplyr workflows
tibble::as_tibble(ap) |> head()
```

## Subsetting and Slicing

```{r series-slice}
# Head / tail
milt_head(ap, 6)
milt_tail(ap, 6)
```

```{r series-window}
# Extract a date window
milt_window(ap,
  start = as.Date("1955-01-01"),
  end   = as.Date("1957-12-01")
)
```

```{r series-bracket}
# [ operator — same as milt_window() but by position
ap[1:24]
```

## Train / Test Split

```{r series-split}
# 80% train, 20% test
spl <- milt_split(ap, ratio = 0.8)
cat("Train:", spl$train$n_timesteps(), "obs\n")
cat("Test: ", spl$test$n_timesteps(),  "obs\n")
```

```{r series-split-at}
# Split at a specific date
spl2 <- milt_split_at(ap, time = as.Date("1958-01-01"))
cat("Train ends:", format(spl2$train$end_time()), "\n")
cat("Test starts:", format(spl2$test$start_time()), "\n")
```

## Concatenation and Resampling

```{r series-concat}
s1 <- ap[1:60]
s2 <- ap[61:144]
recombined <- milt_concat(s1, s2)
recombined$n_timesteps()
```

```{r series-resample}
# Resample to quarterly (from monthly)
ap_q <- milt_resample(ap, period = "quarterly", agg_fn = sum)
print(ap_q)
```

## Handling Missing Values

```{r series-fill-gaps}
# Introduce a gap then fill it
vals <- ap$values()
vals[c(10, 11, 12)] <- NA
s_na <- milt_series(vals, frequency = 12, start = c(1949, 1))

# Fill using linear interpolation
s_filled <- milt_fill_gaps(s_na, method = "linear")
print(s_filled)
```

## Visualising a Series

```{r series-plot}
plot(ap)
```

------------------------------------------------------------------------

# TimeSeries Operations {#sec-tsops}

Beyond creation and slicing (@sec-series), milt supports the same everyday `TimeSeries` operations Python's Darts offers: arithmetic, stacking, generic mapping, and enriching a series with new calendar/holiday components.

## Arithmetic Operations

`+`, `-`, `*`, `/`, and `^` work directly on `MiltSeries` objects — against a scalar, or against another `MiltSeries` that shares an identical time index.

```{r arithmetic}
ap <- milt_series(AirPassengers)

bumped   <- ap * 1.05          # scale every value up 5%
shifted  <- ap + 10            # scalar addition
identity <- ap - ap            # elementwise difference of two series

print(bumped)
identical(as.numeric((ap - ap)$values()), rep(0, ap$n_timesteps()))
```

Series must share an identical time index to be combined with each other — mismatched indices raise a `milt_error_incompatible_series` error rather than silently misaligning data.

## Stacking

`milt_stack()` combines two or more series that share the same time index into a single **multivariate** series, placing each input's components side by side. This is different from `milt_concat()`, which joins series end-to-end along *time* rather than side-by-side along *components*.

```{r stack}
smoothed <- milt_filter(ap, "moving_average", window = 12L)
stacked  <- milt_stack(ap, smoothed)

stacked$n_components()   # 2 — "value" and "value_2"
stacked$is_univariate()  # FALSE
```

## Mapping — On Values, and On Both Timestamps and Values

`milt_step_map()` applies an arbitrary function elementwise. Supply a one-argument function to transform values only, or a **two-argument** function `function(time, value)` to also use the timestamp — milt detects the arity automatically.

```{r map-values}
# Values only
log_out <- milt_step_map(ap, fn = log1p, inverse_fn = expm1)
log_out$series

# Both timestamps and values: zero out the first year on record
trimmed_out <- milt_step_map(ap, fn = function(time, value) {
  ifelse(lubridate::year(time) == lubridate::year(min(time)), 0, value)
})
trimmed_out$series
```

## Adding a Datetime Attribute as an Extra Dimension

`milt_add_datetime_component()` extracts a calendar attribute (month, quarter, day of week, ...) from the time index and appends it as a **new component** — turning a univariate series multivariate, ready to use as a model covariate.

```{r datetime-component}
with_month <- milt_add_datetime_component(ap, "month")
with_month$as_tibble()

# Cyclic encoding avoids the false "December is far from January" discontinuity
with_month_cyclic <- milt_add_datetime_component(ap, "month", cyclic = TRUE)
with_month_cyclic$n_components()   # 3: value, month_sin, month_cos
```

## Adding a Binary Holidays Component

`milt_add_holidays()` appends a `0`/`1` column flagging holiday dates — either from a built-in calendar (currently `"US"` federal holidays) or a custom `dates` vector for any other calendar.

```{r holidays}
with_holidays <- milt_add_holidays(ap, country = "US")
sum(with_holidays$as_tibble()$is_holiday)   # number of holiday months on record

# Custom calendar
custom <- milt_add_holidays(ap, dates = as.Date(c("1955-11-24", "1956-11-22")),
                             name = "is_black_friday_month")
```

## Differencing

Covered in full under *Individual Feature Steps* (@sec-steps) — `milt_step_diff()` applies sequential lagged differencing (e.g. `lags = c(1, 12)` removes a trend, then a yearly seasonal pattern) and is invertible via `milt_step_undiff()`.

## Working with Covariates

Beyond the series' own values, milt supports three kinds of external covariate data via `milt_add_covariates()` — mirroring Darts' `past_covariates` / `future_covariates` / static-covariates model:

-   **`"past"`** — time-varying covariates observed over the same time range as the series (e.g. a promotion flag, competitor pricing). Used for in-sample feature enrichment.
-   **`"future"`** — time-varying covariates available *beyond* the series end (e.g. a calendar feature, a known future price). Used by models to produce covariate-informed forecasts.
-   **`"static"`** — scalar attributes per group that never change over time (e.g. store size, region). Only meaningful for a multi-series (grouped) `MiltSeries`.

`milt_add_covariates()` attaches the data and returns the (mutated) series; `milt_get_covariates()` retrieves what's stored.

```{r covariates-past}
ap    <- milt_series(AirPassengers)
dates <- ap$as_tibble()$time

# A past covariate: the calendar month as a plain integer
month_cov <- data.frame(time = dates, month_num = as.integer(format(dates, "%m")))
ap2 <- milt_add_covariates(ap, month_cov, type = "past", time_col = "time")

milt_get_covariates(ap2, type = "past")
```

Static covariates require a grouped series — one row per group value:

```{r covariates-static}
tbl <- data.frame(
  date  = rep(seq(as.Date("2020-01-01"), by = "month", length.out = 24), 2),
  sales = c(cumsum(rnorm(24, 5)), cumsum(rnorm(24, 8))),
  store = rep(c("A", "B"), each = 24)
)
grouped <- milt_series(tbl, time_col = "date", value_cols = "sales", group_col = "store")

store_attrs <- data.frame(store = c("A", "B"), size_sqft = c(1200, 3400))
grouped2 <- milt_add_covariates(grouped, store_attrs, type = "static", group_col = "store")

milt_get_covariates(grouped2, type = "static")
```

Static covariates are automatically sliced per group and passed through when a model trains group-by-group — see `milt_local_model()` and `milt_global_model()` in *Multi-Series Analysis* (@sec-multi-series).

------------------------------------------------------------------------

# Exploratory Data Analysis {#sec-eda}

`milt_eda()` gives you a quick statistical profile of your series — trend direction, seasonality, stationarity, and outlier counts.

```{r eda}
ap <- milt_series(AirPassengers)
eda_result <- milt_eda(ap)
print(eda_result)
```

## ACF and Decomposition Plots

```{r plot-acf}
milt_plot_acf(ap, lag.max = 36L)
```

```{r plot-decomp}
milt_plot_decomp(ap)
```

## Series Diagnostics with Recommendations

`milt_diagnose()` runs a suite of statistical tests and prints model recommendations.

```{r diagnose}
milt_diagnose(ap)
```

## Inspecting Seasonality

`milt_check_seasonality()` searches the ACF for the strongest *significant* local maximum (via Bartlett's formula for the ACF's standard error under a white-noise null) and reports it as the detected seasonal period — useful for confirming a `period`/`season` argument before fitting a seasonal model.

```{r check-seasonality}
milt_check_seasonality(ap, max_lag = 24L)

# A non-seasonal series correctly reports is_seasonal = FALSE
white_noise <- milt_series(rnorm(120), frequency = "monthly", start = c(2015, 1))
milt_check_seasonality(white_noise, max_lag = 24L)
```

------------------------------------------------------------------------

# Forecasting {#sec-forecasting}

## The Universal Forecasting Interface

milt uses the same three-step pipeline for every model:

```         
milt_model("<name>")  →  milt_fit(<series>)  →  milt_forecast(horizon = <h>)
```

Let's see what models are available:

```{r list-models}
list_milt_models()
```

## Refitting on New Data

In production, new observations keep arriving and a deployed model needs to keep up. `milt_refit()` retrains a model with the **same hyperparameters** it was originally configured with, just on an updated series — no re-tuning, no need to reconstruct the `milt_model(...)` call from scratch:

```{r refit}
initial <- milt_series(AirPassengers[1:120], frequency = 12, start = c(1949, 1))
m       <- milt_model("naive") |> milt_fit(initial)

# A month later, more data has arrived
updated <- milt_series(AirPassengers[1:132], frequency = 12, start = c(1949, 1))
m       <- milt_refit(m, updated)
milt_forecast(m, horizon = 12)
```

## Classical Statistical Models

### ARIMA

Auto-ARIMA automatically selects the best (p, d, q) order.

```{r arima}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)

arima_fit <- milt_model("auto_arima") |> milt_fit(ap)
arima_fct <- milt_forecast(arima_fit, horizon = 24)

print(arima_fct)
plot(arima_fct)
```

### ETS (Exponential Smoothing)

```{r ets}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ets_fct <- milt_model("ets") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 24)

plot(ets_fct)
```

### STL Decomposition

```{r stl}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
stl_fct <- milt_model("stl") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 24)

plot(stl_fct)
```

### Theta Method

```{r theta}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
theta_fct <- milt_model("theta") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(theta_fct)
```

### TBATS

Good for complex seasonality (e.g. daily data with weekly + annual cycles).

```{r tbats}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
tbats_fct <- milt_model("tbats") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(tbats_fct)
```

### Prophet (Facebook)

```{r prophet}
#| eval: !expr requireNamespace("prophet", quietly = TRUE)
prophet_fct <- milt_model("prophet") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 24)

plot(prophet_fct)
```

### Croston (Intermittent Demand)

Croston's method is designed for intermittent (sparse) demand series.

```{r croston}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Simulate intermittent demand
demand <- c(0,0,3,0,0,0,2,0,5,0,0,1,0,0,0,4,0,0,2,0,
            0,0,1,0,3,0,0,0,2,0,0,1,0,0,5,0)
demand_series <- milt_series(demand, frequency = 12, start = c(2020, 1))

croston_fct <- milt_model("croston") |>
  milt_fit(demand_series) |>
  milt_forecast(horizon = 12)

print(croston_fct)
```

### Naive Baseline

Repeats the last observed value for every step of the forecast horizon — equivalent to a random-walk forecast. Always a good baseline to beat; no model deserves trust until it clears this bar.

```{r naive}
naive_fct <- milt_model("naive") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(naive_fct)
```

### Seasonal Naive

Repeats the value from exactly one seasonal cycle ago (for monthly data, the same month from the previous year). A much stronger baseline than plain naive whenever the series has clear seasonality.

```{r snaive}
snaive_fct <- milt_model("snaive") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(snaive_fct)
```

### Drift

Extrapolates a straight line through the first and last training observations — captures a linear trend with no seasonality.

```{r drift}
drift_fct <- milt_model("drift") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(drift_fct)
```

### Mean

Repeats the training-series mean for every forecast step, with a *constant* (non-widening) prediction interval — appropriate only when the series has no trend or seasonality and residuals are i.i.d.

```{r mean}
mean_fct <- milt_model("mean") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(mean_fct)
```

### Moving Average

Forecasts each step as the mean of the last `window` observations, then slides that window forward autoregressively using its *own* forecasts — so far-ahead steps are means of prior forecasts, not just raw data.

```{r moving-average}
ma_fct <- milt_model("moving_average", window = 5L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(ma_fct)
```

## Machine Learning Models

ML models automatically build lag features from the series to train a regression model. You can tune how many lags to use.

### XGBoost

```{r xgboost}
#| eval: !expr requireNamespace("xgboost", quietly = TRUE)
xgb_fct <- milt_model("xgboost", lags = 1:24) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

plot(xgb_fct)
```

### LightGBM

```{r lightgbm}
#| eval: !expr requireNamespace("lightgbm", quietly = TRUE)
lgb_fct <- milt_model("lightgbm", lags = 1:24) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(lgb_fct)
```

### Random Forest

```{r random-forest}
#| eval: !expr requireNamespace("ranger", quietly = TRUE)
rf_fct <- milt_model("random_forest", lags = 1:12) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

plot(rf_fct)
```

### Elastic Net

```{r elastic-net}
#| eval: !expr requireNamespace("glmnet", quietly = TRUE)
en_fct <- milt_model("elastic_net", lags = 1:12) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(en_fct)
```

### Support Vector Machine

```{r svm}
#| eval: !expr requireNamespace("e1071", quietly = TRUE)
svm_fct <- milt_model("svm", lags = 1:12) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(svm_fct)
```

### K-Nearest Neighbours

```{r knn}
knn_fct <- milt_model("knn", lags = 1:12) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

print(knn_fct)
```

## Deep Learning Models (torch)

Deep learning models require the `torch` package with its Lantern C++ backend installed.

``` r
# One-time setup — downloads the C++ runtime
install.packages("torch")
torch::install_torch()
```

More generally, `milt_install_backends()` installs (or upgrades) any set of milt's optional backend packages in one call — `torch` included — so you don't have to hunt down each Suggests dependency by hand:

``` r
milt_install_backends("torch")           # just the DL backend
milt_install_backends(c("xgboost", "prophet"))
milt_install_backends("all")             # every optional backend milt knows about
```

All DL backends follow the same pattern:

``` r
#| eval: false
# N-BEATS
nbeats_fct <- milt_model("nbeats", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# N-HiTS
nhits_fct <- milt_model("nhits", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# Temporal Convolutional Network (TCN)
tcn_fct <- milt_model("tcn", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# Temporal Fusion Transformer (TFT)
tft_fct <- milt_model("tft", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# PatchTST
patch_fct <- milt_model("patch_tst", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# DeepAR (LSTM + Gaussian likelihood — probabilistic by default)
deepar_fct <- milt_model("deepar", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12, num_samples = 100L)

# NLinear (Zeng et al. 2022) — subtracts the last observed value before a
# single linear layer, then adds it back; a deceptively strong, cheap
# baseline that often matches much larger Transformer models
nlinear_fct <- milt_model("nlinear", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# DLinear (Zeng et al. 2022) — decomposes the input window into a
# moving-average trend and a seasonal/residual component, maps each through
# its own linear layer, and sums the two projections
dlinear_fct <- milt_model("dlinear", input_chunk_length = 24L, output_chunk_length = 12L,
                          kernel_size = 25L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# TiDE (Das et al. 2023) — an MLP encoder-decoder built from residual
# blocks (Linear -> ReLU -> Linear -> Dropout + skip connection); a cheap,
# strong alternative to attention-based long-horizon models
tide_fct <- milt_model("tide", input_chunk_length = 24L, output_chunk_length = 12L,
                       hidden_size = 64L, n_encoder_layers = 2L, n_decoder_layers = 2L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

# TSMixer (Chen et al. 2023) — a stack of residual "time-mixing" MLP
# blocks over the lookback window, followed by a linear projection head
# (simplified to time-mixing only, since feature-mixing is degenerate for
# milt's single-channel series)
tsmixer_fct <- milt_model("tsmixer", input_chunk_length = 24L, output_chunk_length = 12L,
                          hidden_size = 64L, n_blocks = 2L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)
```

### Checking the torch device

``` r
milt_torch_device()  # returns cuda if GPU available, else cpu
```

## Darts Bridge Backends (Python, via `reticulate`)

For models not yet natively ported to torch-for-R, milt can transparently delegate to Python's [Darts](https://github.com/unit8co/darts) library through `reticulate`. Call `milt_setup_darts()` once per session to verify (and optionally install) the Python environment.

``` r
milt_setup_darts()                      # check only
milt_setup_darts(install = TRUE)        # check + install darts via pip

rnn_fct <- milt_model("darts_rnn", input_chunk_length = 24L, output_chunk_length = 12L) |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

transformer_fct <- milt_model("darts_transformer") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)

darts_nbeats_fct <- milt_model("darts_nbeats") |>
  milt_fit(ap) |>
  milt_forecast(horizon = 12)
```

------------------------------------------------------------------------

# Filtering Models {#sec-filtering}

Filtering is a different job from forecasting: it estimates the underlying signal from noisy observations at the **same** timestamps as the input — it does not extrapolate into the future. Use `milt_filter()` when you want to denoise an already-observed series, e.g. before plotting, before feature engineering, or before feeding a cleaner signal into a downstream model.

All three filters below are implemented on base R (`stats`) — no additional package dependencies.

## Moving Average Filter

A simple rolling-mean smoother.

```{r filter-ma}
noisy <- milt_series(as.numeric(AirPassengers) + rnorm(144, sd = 15),
                      frequency = "monthly", start = c(1949, 1))

smoothed_ma <- milt_filter(noisy, "moving_average", window = 5L)
plot(smoothed_ma)
```

## Kalman Filter

State-space smoothing via `stats::StructTS()` — models the series as a noisy observation of a hidden local-level (or local-trend / BSM) process and returns the Kalman-smoothed estimate of that hidden state.

```{r filter-kalman}
smoothed_kalman <- milt_filter(noisy, "kalman", type = "level")
plot(smoothed_kalman)
```

## Gaussian Process Filter

Fits a Gaussian Process with a squared-exponential (RBF) kernel to the observed points and returns the posterior mean at those same timestamps — a flexible non-parametric smoother.

```{r filter-gp}
smoothed_gp <- milt_filter(noisy, "gp", length_scale = 3, noise = 5)
plot(smoothed_gp)
```

------------------------------------------------------------------------

# Working with Forecasts {#sec-forecast-objects}

A `MiltForecast` object stores the full forecast result — point estimates, prediction intervals, and (for probabilistic models) sample paths.

```{r forecast-object}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)
fct <- milt_model("ets") |> milt_fit(ap) |> milt_forecast(horizon = 24)

# Inspect
print(fct)
plot(fct)
```

```{r forecast-to-tibble}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Extract as a tibble for further processing
tibble::as_tibble(fct) |> head()
```

## In-sample Predictions and Residuals

```{r predict-residuals}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
m <- milt_model("ets") |> milt_fit(ap)

fitted_vals <- milt_predict(m)
residuals   <- milt_residuals(m)

plot(residuals, type = "l", main = "Residuals", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```

------------------------------------------------------------------------

# Accuracy Metrics {#sec-metrics}

milt ships a comprehensive suite of forecast accuracy functions.

```{r metrics-setup}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap   <- milt_series(AirPassengers)
spl  <- milt_split(ap, ratio = 0.8)
fct  <- milt_model("ets") |> milt_fit(spl$train) |> milt_forecast(spl$test$n_timesteps())

actual <- spl$test$values()
pred   <- tibble::as_tibble(fct)$.mean
```

```{r metrics-all}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Comprehensive accuracy report
milt_accuracy(actual, pred)
```

```{r metrics-individual}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Individual metrics
cat("MAE:  ", milt_mae(actual, pred), "\n")
cat("RMSE: ", milt_rmse(actual, pred), "\n")
cat("MAPE: ", milt_mape(actual, pred), "%\n")
cat("sMAPE:", milt_smape(actual, pred), "%\n")
cat("MASE: ", milt_mase(actual, pred, spl$train$values()), "\n")
cat("R²:   ", milt_r_squared(actual, pred), "\n")
```

### Additional Scale-Free Metrics

Useful when comparing accuracy across series with very different scales, or when `milt_mape()` would blow up on near-zero actuals.

```{r metrics-additional}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
cat("wMAPE (robust to near-zero actuals):", milt_wmape(actual, pred), "\n")
cat("OPE (error of the totals, not per-point):", milt_ope(actual, pred), "\n")
cat("Coefficient of Variation (RMSE / mean):", milt_coefficient_of_variation(actual, pred), "\n")
cat("MARRE (MAE / range of actual):", milt_marre(actual, pred), "\n")
cat("RMSLE (penalises relative, not absolute, error):", milt_rmsle(actual, pred), "\n")
```

`milt_accuracy()` includes `wMAPE`, `OPE`, `CV`, and `MARRE` by default; `RMSLE` is opt-in only (via `metrics = "RMSLE"`) since it errors on values at or below -1:

```{r metrics-accuracy-explicit}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
milt_accuracy(actual, pred, metrics = "RMSLE")
```

## Probabilistic Metrics

```{r probabilistic-metrics}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# For probabilistic forecasts — need sample paths
# (Using simulated samples here for illustration)
samples <- matrix(rnorm(length(actual) * 200, mean = pred, sd = 15),
                  nrow = length(actual), ncol = 200)

cat("CRPS:    ", milt_crps(actual, samples), "\n")
cat("Coverage:", milt_coverage(actual, pred - 1.96*15, pred + 1.96*15), "\n")
```

------------------------------------------------------------------------

# Backtesting and Model Evaluation {#sec-backtest}

## Simple Backtest

```{r backtest}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)

bt <- milt_backtest(
  milt_model("ets"),
  series         = ap,
  initial_window = 84,
  horizon        = 12,
  stride         = 12L
)

print(bt)
tibble::as_tibble(bt) |> head()
```

## Cross-Validation

```{r crossval}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
cv_result <- milt_cv(
  milt_model("auto_arima"),
  series         = ap,
  initial_window = 72,
  horizon        = 12
)

print(cv_result)
```

## Comparing Multiple Models

```{r compare}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
cmp <- milt_compare(
  models = list(
    arima = milt_model("auto_arima"),
    ets   = milt_model("ets"),
    naive = milt_model("naive")
  ),
  series  = ap,
  horizon = 12
)

print(cmp)
plot(cmp)
```

`cmp` is a `MiltComparison` object (backed by the `MiltComparisonR6` class internally) — like every milt result object it has its own `print()`, `plot()`, and `as_tibble()` methods, so `tibble::as_tibble(cmp)` gets you the ranked comparison table for further `dplyr` work.

## Searching for Hyperparameters via Backtesting

`milt_grid_search()` backtests every combination of a parameter grid and ranks them by a chosen metric — works with *any* registered model, not just one specific method. (Darts' own quickstart famously demonstrates this with the Theta method's numeric `theta` parameter; R's classical `forecast::thetaf()` wrapper has no equivalent tunable parameter, so the example below sweeps `moving_average`'s `window` instead — the workflow generalises to any model with a tunable argument, e.g. `xgboost`'s `nrounds`/`max_depth`, or `ets`'s `damped`.)

```{r grid-search}
milt_grid_search(
  "moving_average",
  param_grid = list(window = c(2L, 3L, 5L, 8L, 12L)),
  series     = ap,
  horizon    = 6,
  initial_window = 120L,
  stride         = 6L
)
```

Rows are returned best-first by the chosen metric — the top row is the recommended hyperparameter setting.

## Conformal Prediction: Model-Agnostic Intervals

`milt_conformal()` calibrates prediction intervals around *any* fitted model's forecast — even models with no native probabilistic output — using split-conformal prediction: it backtests the model on walk-forward calibration folds, then uses the empirical quantiles of the resulting per-horizon-step absolute errors as a distribution-free `point ± margin` interval.

```{r conformal}
conformal_fct <- milt_conformal(
  milt_model("naive"), ap, horizon = 12,
  initial_window = 120L, stride = 12L
)

print(conformal_fct)
plot(conformal_fct)
```

------------------------------------------------------------------------

# Ensemble Models {#sec-ensemble}

Combine multiple models with averaging or weighted combinations.

```{r ensemble}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)

m1 <- milt_model("auto_arima") |> milt_fit(ap)
m2 <- milt_model("ets")        |> milt_fit(ap)
m3 <- milt_model("naive")      |> milt_fit(ap)

ens <- milt_ensemble(
  list(arima = m1, ets = m2, naive = m3),
  weights = c(0.5, 0.4, 0.1)
)

ens_fct <- milt_forecast(ens, horizon = 12)
plot(ens_fct)
```

------------------------------------------------------------------------

# Feature Engineering with Pipelines {#sec-pipelines}

The pipeline system lets you compose data transformations into a reusable, fittable object. Pipelines fit on training data and can be applied to new data — preventing data leakage.

## Building a Pipeline

A pipeline always needs a model attached as its final step. Use `milt_pipe_model()` to attach one.

```{r pipeline-build}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap <- milt_series(AirPassengers)

pipe <- milt_pipeline() |>
  milt_pipe_step_lag(lags = 1:12) |>
  milt_pipe_step_rolling(windows = c(3, 6, 12), fns = c("mean", "sd")) |>
  milt_pipe_step_fourier(K = 3) |>
  milt_pipe_step_calendar() |>
  milt_pipe_step_scale() |>
  milt_pipe_model(milt_model("naive"))

print(pipe)
```

## Fitting and Transforming

```{r pipeline-fit}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
spl         <- milt_split(ap, ratio = 0.8)
fitted_pipe <- milt_pipeline_fit(pipe, spl$train)

# Forecast using the fitted pipeline
pipe_fct <- milt_pipeline_forecast(fitted_pipe, horizon = 12)
print(pipe_fct)
```

Sometimes you only want the engineered *features* — not a forecast. `milt_pipeline_transform()` applies a fitted pipeline's transform steps (everything except the model) to any series and hands back the resulting feature table, useful for feeding milt's feature engineering into an external modelling workflow:

```{r pipeline-transform}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
transformed_test <- milt_pipeline_transform(fitted_pipe, spl$test)
head(transformed_test)
```

## Pipeline + Model

Attach a model as the final step of the pipeline:

```{r pipe-model}
#| eval: !expr requireNamespace("xgboost", quietly = TRUE)
full_pipe <- milt_pipeline() |>
  milt_pipe_step_lag(lags = 1:12) |>
  milt_pipe_step_rolling(windows = c(3, 6), fns = "mean") |>
  milt_pipe_step_scale() |>
  milt_pipe_model(milt_model("xgboost"))

pipe_fit <- milt_pipeline_fit(full_pipe, spl$train)
pipe_fct <- milt_pipeline_forecast(pipe_fit, horizon = 12)
plot(pipe_fct)
```

------------------------------------------------------------------------

# Individual Feature Steps {#sec-steps}

The `milt_step_*` functions work directly on a `MiltSeries` without a full pipeline.

```{r step-scale}
ap <- milt_series(AirPassengers)

# Scale the series (z-score normalisation)
scaled <- milt_step_scale(ap)
print(scaled)
```

```{r step-lag}
# Add lag features (returns tibble)
lag_df <- milt_step_lag(ap, lags = c(1, 3, 12))
head(lag_df)
```

```{r step-rolling}
# Rolling statistics
roll_df <- milt_step_rolling(ap, windows = 6, fns = "mean")
head(roll_df)
```

```{r step-fourier}
# Fourier terms for seasonality
fourier_df <- milt_step_fourier(ap, K = 2)
head(fourier_df)
```

```{r step-calendar}
# Calendar features (month, quarter, weekday, …)
cal_df <- milt_step_calendar(ap)
head(cal_df)
```

```{r step-boxcox}
# Box-Cox power transform (auto-estimated lambda) — stabilises variance
bc_out <- milt_step_boxcox(ap)
bc_out$step$lambda()

# Invert back to the original scale
restored <- bc_out$step$inverse_transform(bc_out$series)
all.equal(as.numeric(restored$values()), as.numeric(ap$values()))
```

```{r step-diff}
# Sequential lagged differencing: remove a trend, then yearly seasonality
diff_out <- milt_step_diff(ap, lags = c(1, 12))
diff_out$series$n_timesteps()   # 144 - 13 = 131

# Fully invertible, including reconstructing the dropped leading timestamps
undiffed <- diff_out$step$inverse_transform(diff_out$series)
undiffed$n_timesteps()          # back to 144
```

```{r step-map}
# Generic elementwise transform, with an optional invertible pair
map_out <- milt_step_map(ap, fn = log1p, inverse_fn = expm1)
map_out$series
```

------------------------------------------------------------------------

# Anomaly Detection {#sec-anomaly}

## The Detector Interface

Like models, detectors follow a two-step interface:

```         
milt_detector("<method>")  →  milt_detect(<series>)
```

## Statistical Methods

```{r detect-iqr}
ap <- milt_series(AirPassengers)

# IQR-based detection (classical box-plot rule)
iqr_result <- milt_detector("iqr") |> milt_detect(ap)
print(iqr_result)
plot(iqr_result)
```

```{r detect-stl}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# STL decomposition — flags residuals outside ±3 SD
stl_result <- milt_detector("stl") |> milt_detect(ap)
print(stl_result)
```

```{r detect-grubbs}
#| eval: !expr requireNamespace("tseries", quietly = TRUE)
# Grubbs test (univariate extreme value)
grubbs_result <- milt_detector("grubbs") |> milt_detect(ap)
print(grubbs_result)
```

```{r detect-gesd}
# Generalised ESD test (multiple outliers)
gesd_result <- milt_detector("gesd") |> milt_detect(ap)
print(gesd_result)
```

## Machine Learning Methods

```{r detect-lof}
#| eval: !expr requireNamespace("dbscan", quietly = TRUE)
# Local Outlier Factor
lof_result <- milt_detector("lof") |> milt_detect(ap)
print(lof_result)
```

```{r detect-iforest}
#| eval: !expr requireNamespace("isotree", quietly = TRUE)
# Isolation Forest
iforest_result <- milt_detector("iforest") |> milt_detect(ap)
print(iforest_result)
```

```{r detect-autoencoder}
#| eval: !expr requireNamespace("torch", quietly = TRUE)
# Autoencoder (deep learning)
ae_result <- milt_detector("autoencoder") |> milt_detect(ap)
print(ae_result)
```

## Ensemble Detection

Combine multiple detectors — a point is anomalous only if a majority vote says so.

```{r detect-ensemble}
ens_result <- milt_detector("ensemble",
  detectors = list(
    milt_detector("iqr"),
    milt_detector("gesd")
  )
) |> milt_detect(ap)
print(ens_result)
```

## MiltAnomalies Result Object

```{r anomalies-object}
result <- milt_detector("iqr") |> milt_detect(ap)

# Convert to tibble
tibble::as_tibble(result) |> dplyr::filter(is_anomaly) |> head()
```

## Anomaly Detection via Forecast Residuals

The detectors above look for outliers directly in the raw series. `milt_anomaly_model()` takes a different, complementary approach — the same one Darts calls a `ForecastingAnomalyModel`: wrap **any** already-registered forecaster, fit it, score its residuals, and flag the points whose residual score crosses a threshold. A point the series *looks* normal at but that the model badly mispredicted is exactly the kind of anomaly this catches and a raw-value detector would miss.

`milt_anomaly_score()` supplies the scoring function — `"norm"` (absolute residual), `"difference"` (signed residual), or `"nll_gaussian"` (negative log-likelihood under a Gaussian fit to the residual distribution, penalising extreme residuals more sharply) — and the detector thresholds those scores either by `"quantile"` or explicit `"threshold"` bounds.

```{r anomaly-model}
am    <- milt_anomaly_model(milt_model("naive"), scorer = "norm", detector = "quantile", quantile = 0.95) |>
  milt_fit(ap)
anoms <- milt_detect_anomalies(am)

print(anoms)
plot(anoms)
```

Fitting a `MiltAnomalyModel` clones the wrapped model internally, so the original `milt_model("naive")` object you passed in is never mutated. The wrapped model stays fully accessible — `am$model()` returns the fitted forecaster itself, and `am$scores()` returns the raw per-point anomaly scores that were thresholded — and the result is a standard `MiltAnomalies` object, so every `plot()`/`print()`/`as_tibble()` method you already know keeps working.

------------------------------------------------------------------------

# Change-Point Detection {#sec-changepoints}

```{r changepoints}
#| eval: !expr requireNamespace("changepoint", quietly = TRUE)
ap <- milt_series(AirPassengers)

cp_result <- milt_changepoints(ap, method = "pelt")
print(cp_result)
plot(cp_result)
```

------------------------------------------------------------------------

# Multi-Series Analysis {#sec-multi-series}

## Creating a Multi-Series Object

```{r multi-series-create}
# Simulate three store sales series
set.seed(42)
sales_df <- data.frame(
  date  = rep(seq(as.Date("2019-01-01"), by = "month", length.out = 48), 3),
  store = rep(c("A", "B", "C"), each = 48),
  sales = c(
    100 + cumsum(rnorm(48, 1, 3)),
    200 + cumsum(rnorm(48, 2, 5)),
    150 + cumsum(rnorm(48, 0.5, 2))
  )
)

ms <- milt_series(sales_df,
  time_col  = "date",
  value_col = "sales",
  group_col = "store",
  frequency = 12
)

print(ms)
```

## Fitting Models Across All Groups

```{r multi-fit}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
# Fit ETS to every store independently
ms_fit <- milt_model("ets") |> milt_fit(ms)
ms_fct <- milt_forecast(ms_fit, horizon = 12)
print(ms_fct)
```

## Global Models — One Shared Model Across All Groups

`milt_local_model()` (above) fits one **independent** model per group. `milt_global_model()` takes the opposite approach: it trains a single **shared** model (`"knn"` or `"xgboost"`) on every group's data pooled together, one-hot encoding group identity and appending any static covariates as constant-per-group feature columns — the same idea as Darts training one model across a `Sequence[TimeSeries]`.

Global models are worth reaching for when groups are too short individually to train a good local model on their own, but collectively share enough structure that pooling helps — the shared model can borrow statistical strength across stores/sensors/customers.

```{r global-model}
gm <- milt_global_model("knn", lags = 1:6, k = 3L) |> milt_fit(ms)

# forecast() targets one group at a time
gm$forecast(horizon = 6, group = "A")

# forecast_all() runs every group through the one shared model at once
all_fcts <- gm$forecast_all(horizon = 6)
all_fcts[["B"]]
```

Despite sharing one set of learned weights, each group still gets a genuinely different forecast — the group indicator (and any static covariates) let the shared model tell the groups apart.

------------------------------------------------------------------------

# Hierarchical Reconciliation {#sec-hierarchical}

When your series form a hierarchy (e.g., country → region → store), forecasts at lower levels should sum to the forecast at the higher level. `milt_reconcile()` enforces this.

```{r hierarchical}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
set.seed(42)
# Three individual series: total, region_A, region_B
make_hier <- function(n, mean_drift) {
  milt_series(300 + cumsum(rnorm(n, mean_drift, 3)),
              frequency = 12, start = c(2019, 1))
}

s_total    <- make_hier(48, 3)
s_region_a <- make_hier(48, 1)
s_region_b <- make_hier(48, 2)

# Fit ETS independently on each series
fct_total    <- milt_model("ets") |> milt_fit(s_total)    |> milt_forecast(horizon = 12)
fct_region_a <- milt_model("ets") |> milt_fit(s_region_a) |> milt_forecast(horizon = 12)
fct_region_b <- milt_model("ets") |> milt_fit(s_region_b) |> milt_forecast(horizon = 12)

# Summation matrix S: rows = [total, region_A, region_B], cols = [region_A, region_B]
S <- matrix(
  c(1, 1,   # total  = region_A + region_B
    1, 0,   # region_A
    0, 1),  # region_B
  nrow = 3, ncol = 2, byrow = TRUE
)

# Reconcile — forecasts is a named list of MiltForecast objects
reconciled <- milt_reconcile(
  forecasts = list(total = fct_total, region_A = fct_region_a, region_B = fct_region_b),
  S         = S,
  method    = "ols"
)
print(reconciled)
tibble::as_tibble(reconciled)
```

------------------------------------------------------------------------

# Time Series Classification {#sec-classify}

Classify time series into predefined categories. Useful for product type detection, fault classification, customer segmentation, and more.

```{r classify}
#| eval: !expr requireNamespace("ranger", quietly = TRUE)
# Build labelled training data
set.seed(1)
make_series <- function(n, trend, noise) {
  milt_series(cumsum(rnorm(n, trend, noise)),
    frequency = 1, start = c(2020, 1))
}

# Three classes: trending up, flat, trending down
train_series <- list(
  s1 = make_series(50,  0.5, 1),
  s2 = make_series(50,  0.0, 1),
  s3 = make_series(50, -0.5, 1),
  s4 = make_series(50,  0.6, 1),
  s5 = make_series(50, -0.4, 1)
)

labels <- c("up", "flat", "down", "up", "down")

# Fit classifier
clf <- milt_classifier("random_forest") |>
  milt_classify_fit(train_series, labels)

# Predict on new series
new_series <- list(
  t1 = make_series(50,  0.4, 1),
  t2 = make_series(50, -0.3, 1)
)

predictions <- milt_classify_predict(clf, new_series)
print(predictions)
```

------------------------------------------------------------------------

# Time Series Clustering {#sec-cluster}

Group similar time series together without labels.

```{r cluster}
#| eval: !expr requireNamespace("dtw", quietly = TRUE)
# Simulate 9 series from 3 underlying patterns
set.seed(42)
series_list <- lapply(1:9, function(i) {
  group   <- (i - 1) %% 3
  trend   <- c(0.3, 0, -0.3)[group + 1]
  vals    <- cumsum(rnorm(60, trend, 1))
  milt_series(vals, frequency = 12, start = c(2018, 1))
})

cl_result <- milt_cluster(series_list, k = 3, method = "dtw")
print(cl_result)
```

------------------------------------------------------------------------

# Explainability {#sec-explain}

`milt_explain()` uses SHAP-style feature importance to show which lag/feature values drove each forecast.

```{r explain}
#| eval: !expr requireNamespace("xgboost", quietly = TRUE)
ap <- milt_series(AirPassengers)

xgb_model <- milt_model("xgboost", lags = 1:12) |> milt_fit(ap)
explanation <- milt_explain(xgb_model, ap)

print(explanation)
plot(explanation)
```

------------------------------------------------------------------------

# Causal Impact Analysis {#sec-causal}

Measure the effect of an intervention (e.g., a marketing campaign, product launch, policy change) by building a counterfactual forecast of what *would have* happened without it.

```{r causal-impact}
#| eval: !expr requireNamespace("CausalImpact", quietly = TRUE)
set.seed(1)
n        <- 100
pre_n    <- 70
baseline <- cumsum(rnorm(n, 0.5, 1)) + 100

# Add a treatment effect from observation 71 onwards
treatment                       <- baseline
treatment[(pre_n + 1):n] <- treatment[(pre_n + 1):n] + 15

dates  <- seq(as.Date("2016-01-01"), by = "week", length.out = n)
series <- milt_series(treatment, frequency = 52, start = c(2016, 1))

ci_result <- milt_causal_impact(
  series,
  event_time = dates[pre_n + 1]
)

print(ci_result)
plot(ci_result)
```

------------------------------------------------------------------------

# Saving and Loading Models {#sec-save-load}

Trained models can be serialised to disk and reloaded in a fresh session.

```{r save-load}
#| eval: !expr requireNamespace("forecast", quietly = TRUE)
ap    <- milt_series(AirPassengers)
model <- milt_model("ets") |> milt_fit(ap)

# Save to a file
tmp <- tempfile(fileext = ".milt")
milt_save(model, path = tmp)

# Load from file
loaded <- milt_load(tmp)
fct    <- milt_forecast(loaded, horizon = 12)
print(fct)

# Clean up
file.remove(tmp)
```

------------------------------------------------------------------------

# Model Registry {#sec-registry}

milt has a global model registry. You can inspect it and register your own custom model backends.

```{r registry-list}
# Inspect all registered models
models_tbl <- list_milt_models()
print(models_tbl)
```

Two lower-level helpers back `list_milt_models()` and `milt_model()` — useful when you're writing code that needs to check the registry itself, e.g. inside a package extending milt with new backends:

```{r registry-helpers}
is_registered_model("naive")       # TRUE
is_registered_model("not_a_model") # FALSE

get_milt_model_class("naive")      # the MiltNaive R6 generator
```

## Registering a Custom Model

You can register any R6 class that follows the `MiltModelBase` interface:

``` r
# Minimal custom model skeleton
MyCustomModel <- R6::R6Class(
  classname = "MyCustomModel",
  inherit   = MiltModelBase,
  public = list(
    fit = function(series) {
      # ... fit logic ...
      private$.is_fitted <- TRUE
      invisible(self)
    },
    forecast = function(horizon, ...) {
      # ... forecast logic ...
    },
    predict = function(series = NULL, ...) {
      # ... in-sample prediction ...
    },
    residuals = function(...) {
      # ... residuals ...
    }
  )
)

# Register it
register_milt_model(
  name        = "my_model",
  class       = MyCustomModel,
  description = "My custom model"
)

# Use it just like any built-in model
milt_model("my_model") |> milt_fit(ap) |> milt_forecast(horizon = 12)
```

------------------------------------------------------------------------

# Deployment {#sec-deployment}

## Generating a Report

`milt_report()` renders an HTML / PDF report from a fitted forecast.

``` r
# Requires pandoc and rmarkdown
ap    <- milt_series(AirPassengers)
model <- milt_model("ets") |> milt_fit(ap)

milt_report(
  series      = ap,
  models      = list(ets = model),
  horizon     = 24L,
  output_format = "html",
  output_file   = "my_forecast_report.html"
)
```

## Interactive Dashboard

`milt_dashboard()` launches a Shiny dashboard to explore forecasts interactively.

``` r
# Requires shiny
milt_dashboard(
  model  = model,
  series = ap
)
```

## REST API with Plumber

`milt_serve()` wraps a trained model as a REST API using the `plumber` package.

``` r
# Requires plumber
api <- milt_serve(model, port = 8080)

# Then call your model:
# POST http://localhost:8080/predict  {"horizon": 12}
```

------------------------------------------------------------------------

# Deep Dive: The Architecture {#sec-architecture}

Understanding how milt is structured helps when extending it.

## 4-Layer Architecture

```         
┌──────────────────────────────────────────────────┐
│  Layer 1: User API                               │
│  milt_model(), milt_fit(), milt_forecast(), …    │
│  S3 generics: print(), plot(), summary()         │
├──────────────────────────────────────────────────┤
│  Layer 2: Core Engine                            │
│  MiltSeriesR6, MiltForecastR6, ModelRegistry     │
│  MiltAnomaliesR6, MiltBacktestR6, …              │
├──────────────────────────────────────────────────┤
│  Layer 3: Model Backends                         │
│  backend-arima.R, backend-ets.R, backend-xgb.R  │
│  backend-nbeats.R, detect-iqr.R, …              │
├──────────────────────────────────────────────────┤
│  Layer 4: Data Layer                             │
│  tsibble, arrow, data.table, tibble              │
└──────────────────────────────────────────────────┘
```

## Adding a New Model Backend

1.  Create `R/backend-mymodel.R`
2.  Define an R6 class inheriting `MiltModelBase`
3.  Implement `fit()`, `forecast()`, `predict()`, `residuals()`
4.  Call `register_milt_model()` at package load time
5.  Done — no changes needed anywhere else

## The Dependency Strategy

| Type | Packages | Behaviour |
|----|----|----|
| **Imports** | `R6`, `rlang`, `cli`, `dplyr`, `tibble`, `tsibble`, … | Always available |
| **Suggests** | `forecast`, `xgboost`, `torch`, `prophet`, … | Loaded on demand via `check_installed_backend()` |

------------------------------------------------------------------------

# Complete End-to-End Example {#sec-e2e}

Let's walk through a complete forecasting workflow — from raw data to deployed forecast — in one coherent example.

```{r e2e-full}
#| eval: !expr (requireNamespace("forecast", quietly = TRUE) && requireNamespace("xgboost", quietly = TRUE))

# ── 1. Create the series ──────────────────────────────────────────────────────
ap <- milt_series(AirPassengers, frequency = 12, start = c(1949, 1))
print(ap)

# ── 2. Explore ────────────────────────────────────────────────────────────────
eda_result <- milt_eda(ap)
print(eda_result)

# ── 3. Diagnose ───────────────────────────────────────────────────────────────
milt_diagnose(ap)

# ── 4. Split ──────────────────────────────────────────────────────────────────
spl <- milt_split(ap, ratio = 0.8)

# ── 5. Detect anomalies on training data ─────────────────────────────────────
anomalies <- milt_detector("iqr") |> milt_detect(spl$train)
print(anomalies)

# ── 6. Fit and compare three models ──────────────────────────────────────────
cmp <- milt_compare(
  models = list(
    ets   = milt_model("ets"),
    arima = milt_model("auto_arima"),
    knn   = milt_model("knn", lags = 1:12)
  ),
  series  = spl$train,
  horizon = spl$test$n_timesteps()
)
print(cmp)

# ── 7. Pick best and forecast ─────────────────────────────────────────────────
best_model <- milt_model("ets") |> milt_fit(ap)
final_fct  <- milt_forecast(best_model, horizon = 24)

plot(final_fct)

# ── 8. Score on held-out test set ─────────────────────────────────────────────
test_fct <- milt_model("ets") |>
  milt_fit(spl$train) |>
  milt_forecast(spl$test$n_timesteps())

milt_accuracy(spl$test$values(), tibble::as_tibble(test_fct)$.mean)
```

------------------------------------------------------------------------

# Quick Reference {#sec-reference}

## Series Functions

| Function | Purpose |
|----|----|
| `milt_series()` | Create a MiltSeries |
| `milt_head()` / `milt_tail()` | First / last N observations |
| `milt_window()` | Subset by date range |
| `s[i]` | Subset by integer index, logical, or date range |
| `milt_split()` | Train/test split by ratio |
| `milt_split_at()` | Train/test split by date |
| `milt_concat()` | Join two series end-to-end (by time) |
| `milt_stack()` | Combine series side by side (by component) |
| `milt_resample()` | Change frequency |
| `milt_fill_gaps()` | Interpolate missing values |
| `milt_filter()` | Smooth an observed series (moving average / Kalman / GP) |
| `+`, `-`, `*`, `/`, `^` | Arithmetic between series and/or scalars |
| `milt_add_datetime_component()` | Append a calendar attribute as a new component |
| `milt_add_holidays()` | Append a binary holiday indicator component |
| `milt_check_seasonality()` | Detect the strongest significant seasonal period |
| `milt_to_tibble()` | Convert to tibble |
| `milt_to_ts()` | Convert to base R `ts` |
| `milt_to_tsibble()` | Convert to tsibble |
| `as.data.table()` | Convert to data.table |
| `milt_write_parquet()` / `milt_read_parquet()` | Parquet I/O via `arrow` |

## Model Functions

| Function                        | Purpose                               |
|---------------------------------|---------------------------------------|
| `list_milt_models()`            | Show all registered models            |
| `milt_model(name)`              | Instantiate a model                   |
| `milt_fit(model, series)`       | Train the model                       |
| `milt_forecast(model, horizon)` | Generate forecasts                    |
| `milt_predict(model)`           | In-sample fitted values               |
| `milt_residuals(model)`         | Training residuals                    |
| `milt_backtest()`               | Rolling-window backtest               |
| `milt_cv()`                     | Time-series cross-validation          |
| `milt_compare()`                | Side-by-side model comparison         |
| `milt_grid_search()`            | Hyperparameter search via backtesting |
| `milt_conformal()`              | Model-agnostic conformal intervals    |
| `milt_ensemble()`               | Weighted model ensemble               |
| `milt_local_model()`            | Fit any model independently per group |

## Model Names

| Name                  | Type          | Package            |
|-----------------------|---------------|--------------------|
| `"naive"`             | Baseline      | —                  |
| `"snaive"`            | Baseline      | —                  |
| `"drift"`             | Baseline      | —                  |
| `"mean"`              | Baseline      | —                  |
| `"moving_average"`    | Baseline      | —                  |
| `"auto_arima"`        | Classical     | forecast           |
| `"ets"`               | Classical     | forecast           |
| `"stl"`               | Classical     | forecast           |
| `"theta"`             | Classical     | forecast           |
| `"tbats"`             | Classical     | forecast           |
| `"croston"`           | Classical     | forecast           |
| `"prophet"`           | Classical     | prophet            |
| `"xgboost"`           | ML            | xgboost            |
| `"lightgbm"`          | ML            | lightgbm           |
| `"random_forest"`     | ML            | ranger             |
| `"elastic_net"`       | ML            | glmnet             |
| `"svm"`               | ML            | e1071              |
| `"knn"`               | ML            | —                  |
| `"nbeats"`            | Deep Learning | torch              |
| `"nhits"`             | Deep Learning | torch              |
| `"tcn"`               | Deep Learning | torch              |
| `"tft"`               | Deep Learning | torch              |
| `"patch_tst"`         | Deep Learning | torch              |
| `"deepar"`            | Deep Learning | torch              |
| `"nlinear"`           | Deep Learning | torch              |
| `"dlinear"`           | Deep Learning | torch              |
| `"tide"`              | Deep Learning | torch              |
| `"tsmixer"`           | Deep Learning | torch              |
| `"darts_rnn"`         | Deep Learning | reticulate + Darts |
| `"darts_transformer"` | Deep Learning | reticulate + Darts |
| `"darts_nbeats"`      | Deep Learning | reticulate + Darts |

## Filtering Models

| Name               | Method                                        |
|--------------------|-----------------------------------------------|
| `"moving_average"` | Rolling-mean smoother                         |
| `"kalman"`         | State-space smoothing via `stats::StructTS()` |
| `"gp"`             | Gaussian Process smoother (RBF kernel)        |

## Feature / Transform Steps

| Function | Purpose |
|----|----|
| `milt_step_lag()` | Lag features |
| `milt_step_rolling()` | Rolling-window statistics |
| `milt_step_fourier()` | Fourier seasonal terms |
| `milt_step_calendar()` | Calendar dummy features |
| `milt_step_scale()` / `milt_step_unscale()` | Invertible normalisation (z-score/minmax/robust) |
| `milt_step_boxcox()` / `milt_step_unboxcox()` | Invertible Box-Cox power transform |
| `milt_step_diff()` / `milt_step_undiff()` | Invertible sequential lagged differencing |
| `milt_step_map()` | Generic (optionally time-aware, invertible) elementwise transform |

## Anomaly Detectors

| Name            | Method                      |
|-----------------|-----------------------------|
| `"iqr"`         | IQR box-plot rule           |
| `"stl"`         | STL decomposition residuals |
| `"grubbs"`      | Grubbs extreme-value test   |
| `"gesd"`        | Generalised ESD test        |
| `"lof"`         | Local Outlier Factor        |
| `"iforest"`     | Isolation Forest            |
| `"autoencoder"` | Deep learning autoencoder   |
| `"ensemble"`    | Majority-vote combination   |

## Metrics

| Function | Description |
|----|----|
| `milt_accuracy()` | Multiple metrics at once |
| `milt_mae()` | Mean Absolute Error |
| `milt_mse()` | Mean Squared Error |
| `milt_rmse()` | Root Mean Squared Error |
| `milt_mape()` | Mean Absolute % Error |
| `milt_smape()` | Symmetric MAPE |
| `milt_wmape()` | Weighted MAPE (robust to near-zero actuals) |
| `milt_ope()` | Overall Percentage Error (of the totals) |
| `milt_coefficient_of_variation()` | RMSE as a fraction of the mean |
| `milt_marre()` | Mean Absolute Ranged Relative Error |
| `milt_rmsle()` | Root Mean Squared Log Error |
| `milt_mase()` | Mean Abs Scaled Error |
| `milt_rmsse()` | Root Mean Sq Scaled Error |
| `milt_mrae()` | Mean Relative Absolute Error (vs. a benchmark) |
| `milt_r_squared()` | R² |
| `milt_crps()` | Continuous Ranked Probability Score |
| `milt_coverage()` | Prediction interval coverage |
| `milt_pinball()` | Pinball / quantile loss |
| `milt_winkler()` | Winkler interval score |

------------------------------------------------------------------------

# Session Info {#sec-session}

```{r session-info}
sessionInfo()
```
