---
title: "milt Package Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{milt Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(milt)
```

## What milt is

`milt` stands for Modern Integrated Library for Timeseries. The package is
designed to provide a unified R interface for the major families of time series
work:

- forecasting
- feature engineering
- backtesting and model comparison
- anomaly detection and changepoint analysis
- multi-series analysis
- explainability and diagnostics
- deployment and reporting

The package is intentionally broad. Instead of exposing one API for classical
models, another for machine learning, and another for deep learning, it uses a
single workflow pattern across the library.

## Core design principle

The central workflow is:

```r
milt_model("<name>") |> milt_fit(series) |> milt_forecast(horizon = 12)
```

That same structure applies regardless of model family. This makes it easier
to:

- swap models without rewriting the rest of the workflow
- evaluate models under the same backtesting setup
- build ensembles and comparisons across heterogeneous backends
- standardize how outputs are printed, plotted, and converted to tibbles

## Main object model

The package is centered on a small set of reusable objects.

### `MiltSeries`

`MiltSeries` is the canonical container for all time series data inside the
package. Most user-facing functions either consume or return a `MiltSeries`.

Create it from many source types:

```{r create-series}
air <- milt_series(AirPassengers)
print(air)
```

Common accessors:

```{r series-accessors}
air$n_timesteps()
air$freq()
air$start_time()
air$end_time()
head(air$values())
```

### `MiltModel`

`MiltModel` objects represent model specifications before fitting and fitted
model states after training.

```{r model-example}
model <- milt_model("naive")
model
```

### `MiltForecast`

Forecast results are returned as a `MiltForecast`, which contains point
forecasts, interval columns, and plotting/printing helpers.

```{r forecast-example}
fct <- milt_model("naive") |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)

print(fct)
head(fct$as_tibble())
```

### Other result objects

Other workflows return their own result classes, including:

- `MiltBacktest`
- `MiltComparison`
- `MiltAnomalies`
- `MiltExplanation`
- `MiltCausalImpact`
- `MiltChangepoints`

## Forecasting workflows

Forecasting is one major pillar of the package.

```{r forecasting, eval = requireNamespace("forecast", quietly = TRUE)}
fct_ets <- milt_model("ets") |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)

plot(fct_ets)
```

The forecasting layer supports several families of models:

- baseline models such as naive, seasonal naive, and drift
- classical statistical models such as ETS, Auto-ARIMA, Theta, STL, and TBATS
- machine-learning models such as XGBoost, LightGBM, random forest, elastic net, SVM, and related backends
- ensemble workflows
- deep-learning backends through `torch` and `reticulate` integrations

Inspect the available registry:

```{r model-registry}
list_milt_models()
```

## Evaluation and backtesting

The package includes multiple levels of evaluation.

### Train/test splits

```{r split}
spl <- milt_split(air, ratio = 0.8)
spl$train$n_timesteps()
spl$test$n_timesteps()
```

### Accuracy metrics

```{r accuracy}
fct_test <- milt_model("naive") |>
  milt_fit(spl$train) |>
  milt_forecast(spl$test$n_timesteps())

milt_accuracy(
  actual = spl$test$values(),
  predicted = fct_test$as_tibble()$.mean
)
```

### Rolling-origin backtesting

```{r backtest}
bt <- milt_backtest(
  model = milt_model("naive"),
  series = air,
  horizon = 12,
  initial_window = 108L,
  stride = 12L
)

print(bt)
```

### Model comparison

```{r compare}
cmp <- milt_compare(
  models = list(
    naive = milt_model("naive"),
    drift = milt_model("drift")
  ),
  series = air,
  horizon = 12,
  initial_window = 108L,
  stride = 12L
)

print(cmp)
```

## Feature engineering and pipelines

`milt` supports direct feature steps and pipeline composition.

### Direct steps

```{r direct-steps}
air_features <- air |>
  milt_step_lag(lags = 1:6) |>
  milt_step_rolling(windows = 3L, fns = "mean") |>
  milt_step_fourier(period = 12, K = 2)

print(air_features)
```

Available step families include:

- lag features
- rolling statistics
- Fourier seasonality terms
- calendar features
- scaling and inverse scaling

### Pipelines

Pipelines let you define repeatable transformations and model application in a
single object.

```{r pipeline, eval = FALSE}
pipe <- milt_pipeline() |>
  milt_pipe_step_lag(lags = 1:12) |>
  milt_pipe_step_rolling(windows = c(3L, 6L), fns = "mean") |>
  milt_pipe_model("xgboost")

pipe <- milt_pipeline_fit(pipe, air)
fct  <- milt_pipeline_forecast(pipe, horizon = 12)
```

## Diagnostics and analysis

The package includes diagnostics beyond forecasting.

### Series diagnostics

```{r diagnose}
dx <- milt_diagnose(air)
print(dx)
```

### Exploratory analysis

```{r eda}
eda <- milt_eda(air)
print(eda)
```

### Explainability

For supported model types, `milt_explain()` provides feature-importance style
summaries and plots.

```{r explain, eval = FALSE}
exp <- milt_explain(fitted_model)
print(exp)
plot(exp)
```

## Anomaly detection and changepoints

`milt` also includes detection workflows for unusual behavior in series.

```{r anomalies}
detector <- milt_detector("iqr")
anom <- milt_detect(detector, air)
print(anom)
```

Changepoint analysis is exposed separately:

```{r changepoints}
cp <- milt_changepoints(air)
print(cp)
```

## Multi-series workflows

For grouped or panel data, the package includes:

- clustering
- classification
- reconciliation

Those workflows are documented further in the dedicated multi-series vignette,
but they fit into the same object-oriented structure as the rest of the
package.

## Deployment layer

The package includes several operational helpers:

- `milt_save()` and `milt_load()`
- `milt_serve()`
- `milt_dashboard()`
- `milt_report()`

This makes the package useful beyond experimentation. The intended scope covers
development, validation, and handoff or delivery workflows as well.

## Built-in datasets

The package ships with example data for demonstrations and tests:

```{r datasets}
data(milt_air)
data(milt_retail)
data(milt_energy)
```

- `milt_air`: monthly airline passenger data
- `milt_retail`: synthetic multi-series retail data
- `milt_energy`: synthetic energy demand data with covariates

## Documentation map

The pkgdown site should be read in layers:

1. Start with this overview.
2. Read the getting-started guide for the basic workflow.
3. Use the task-focused vignettes for forecasting, anomaly detection,
   pipelines, multi-series work, deep learning, and deployment.
4. Use the reference index for exact function-level documentation.

## Summary

`milt` is not just a small forecasting wrapper. It is structured as a broader
time series workbench:

- one core series type
- one model workflow
- one evaluation pattern
- multiple interchangeable backends
- supporting layers for diagnostics, anomaly detection, multi-series tasks, and deployment

That is the main organizing idea behind the package and the website.
