---
title: "Model Reference Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Reference Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = FALSE
)
library(milt)
```

This page is the complete reference for every forecasting model available in
milt.  Every model uses the same three-step interface:

```{r api, eval = FALSE}
milt_model("<name>", <hyperparameters>) |>
  milt_fit(series) |>
  milt_forecast(horizon = 12)
```

Use `list_milt_models()` to print the full registry at any time.

---

# Choosing a model

| Family | Models | Best for |
|---|---|---|
| Baseline | `naive`, `snaive`, `drift` | Benchmarks, sparse data |
| Classical | `auto_arima`, `ets`, `theta`, `tbats`, `stl` | Univariate, well-behaved seasonality |
| Intermittent | `croston` | Demand series with many zeros |
| Machine learning | `xgboost`, `lightgbm`, `random_forest`, `elastic_net`, `svm`, `knn` | Complex patterns, many covariates |
| Deep learning | `nbeats`, `nhits`, `tft`, `tcn`, `patch_tst`, `deepar` | Long series, large datasets |
| Prophet | `prophet` | Strong multi-seasonality, holidays |

---

# Baseline models

## `naive` — Naïve forecast

Repeats the last observed value for every forecast step.  Equivalent to a
random-walk forecast.  A mandatory benchmark — any useful model should
outperform it.

**No hyperparameters.**  No package dependency.

```{r naive}
air <- milt_series(AirPassengers)

naive_fct <- milt_model("naive") |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)

plot(naive_fct)
```

---

## `snaive` — Seasonal naïve forecast

Repeats the last full seasonal cycle.  For monthly data this copies the same
month from the previous year.  A strong baseline for series with clear
seasonality.

**No package dependency.**

| Parameter | Type | Default | Description |
|---|---|---|---|
| `period` | integer or NULL | `NULL` | Seasonal period. `NULL` auto-detects from series frequency. |

```{r snaive}
snaive_fct <- milt_model("snaive") |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

---

## `drift` — Drift forecast

Extrapolates the average trend observed between the first and last training
observations.

**No hyperparameters.**  No package dependency.

```{r drift}
drift_fct <- milt_model("drift") |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

---

# Classical statistical models

## `auto_arima` — Automatic ARIMA

Wraps `forecast::auto.arima()`.  Searches over ARIMA(p,d,q)(P,D,Q) orders
using AICc or BIC, with optional seasonal differencing and stepwise search.

**Requires:** `forecast`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `stepwise` | logical | `TRUE` | Use stepwise AIC search. `FALSE` tries all combinations (slower, sometimes better). |
| `approximation` | logical or NULL | `NULL` | Use approximations for large series. `NULL` auto-selects based on series length. |
| `seasonal` | logical | `TRUE` | Include a seasonal ARIMA component. Set `FALSE` for clearly non-seasonal data. |
| `...` | — | — | Any argument accepted by `forecast::auto.arima()`. |

```{r auto-arima}
arima_fct <- milt_model("auto_arima",
                         stepwise     = TRUE,
                         seasonal     = TRUE) |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)

plot(arima_fct)
tibble::as_tibble(arima_fct)
```

**When to use:**  A solid default for univariate monthly or quarterly series.
Use `stepwise = FALSE` for a thorough search when fit time is not a concern.

---

## `ets` — Exponential Smoothing (ETS)

Wraps `forecast::ets()`.  Fits an Error-Trend-Seasonality state-space model.
The model string `"ZZZ"` lets the algorithm choose the best combination
automatically.

**Requires:** `forecast`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `model` | character | `"ZZZ"` | ETS model string. Each position is E, T, or S type: `A` (additive), `M` (multiplicative), `N` (none), or `Z` (auto). Examples: `"AAN"` = additive error + additive trend + no season; `"MAM"` = multiplicative error + additive trend + multiplicative season. |
| `damped` | logical or NULL | `NULL` | Force a damped trend? `NULL` auto-selects. |
| `...` | — | — | Any argument accepted by `forecast::ets()`. |

```{r ets}
# Automatic selection
ets_auto <- milt_model("ets") |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)

# Force additive-error multiplicative-trend multiplicative-season (MAM)
ets_mam  <- milt_model("ets", model = "MAM") |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)
```

**When to use:**  Robust for a wide range of univariate series.  Works well
when the variance grows proportionally to the level (use `"ZZZ"` or `"MAM"`).

---

## `theta` — Theta method

Wraps `forecast::thetaf()`.  Decomposes the series into two "theta lines" with
different slopes and combines their forecasts.  Performs well on M3 competition
data.

**Requires:** `forecast`

No hyperparameters (all options are passed through `...` to
`forecast::thetaf()`).

```{r theta}
theta_fct <- milt_model("theta") |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

---

## `stl` — STL decomposition forecast

Wraps `forecast::stlf()`.  Decomposes the series using STL (Seasonal and Trend
decomposition using Loess), forecasts the seasonally-adjusted series with ETS
or ARIMA, then re-seasonalises.

**Requires:** `forecast`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `method` | character | `"ets"` | Model for the remainder: `"ets"` or `"arima"`. |
| `s.window` | character or integer | `"periodic"` | STL seasonal smoothing window. `"periodic"` enforces a constant seasonal pattern; an odd positive integer allows the pattern to evolve. |
| `robust` | logical | `FALSE` | Use robust STL fitting (resistant to outliers). |
| `...` | — | — | Any argument accepted by `forecast::stlf()`. |

```{r stl}
stl_fct <- milt_model("stl",
                       method   = "ets",
                       s.window = "periodic",
                       robust   = TRUE) |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)
```

---

## `tbats` — TBATS

Wraps `forecast::tbats()`.  Handles multiple seasonalities, Box-Cox
transformations, ARMA errors, and damped trends.  Computationally expensive
but powerful for complex seasonal patterns (e.g., hourly data with daily and
weekly seasons).

**Requires:** `forecast`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `use_box_cox` | logical or NULL | `NULL` | Apply Box-Cox transformation. `NULL` auto-selects. |
| `use_trend` | logical or NULL | `NULL` | Include a global trend. `NULL` auto-selects. |
| `use_damped_trend` | logical or NULL | `NULL` | Damp the trend. `NULL` auto-selects. |
| `seasonal_periods` | numeric vector or NULL | `NULL` | Override detected seasonal periods (e.g., `c(7, 365.25)`). |
| `use_arma_errors` | logical | `TRUE` | Include ARMA residual errors. |

```{r tbats}
tbats_fct <- milt_model("tbats",
                          use_box_cox   = NULL,
                          use_trend     = NULL,
                          use_arma_errors = TRUE) |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)
```

---

## `croston` — Croston's method

Wraps `forecast::croston()`.  Designed for **intermittent demand** series —
where many observations are zero with sporadic positive values.  Separately
smooths the non-zero demand sizes and the intervals between them.

**Requires:** `forecast`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `alpha` | numeric in (0, 1] | `0.1` | Smoothing parameter. Smaller values produce smoother forecasts. |

```{r croston}
# Simulated intermittent series
set.seed(42)
demand <- c(0,0,3,0,0,0,5,0,2,0,0,4,0,0,1)
demand_s <- milt_series(
  tibble::tibble(t = seq_along(demand), y = demand),
  time_col = "t", value_cols = "y"
)

croston_fct <- milt_model("croston", alpha = 0.1) |>
  milt_fit(demand_s) |>
  milt_forecast(horizon = 6)
```

---

## `prophet` — Facebook Prophet

Wraps `prophet::prophet()`.  Decomposes the series into trend, seasonality,
and holiday components.  Designed for business time series with strong
multi-seasonality and irregular holidays.

**Requires:** `prophet`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `yearly_seasonality` | logical or `"auto"` | `"auto"` | Include a yearly Fourier seasonal component. |
| `weekly_seasonality` | logical or `"auto"` | `"auto"` | Include a weekly Fourier seasonal component. |
| `daily_seasonality` | logical or `"auto"` | `"auto"` | Include a daily Fourier seasonal component. |
| `seasonality_mode` | character | `"additive"` | `"additive"` or `"multiplicative"`. Use multiplicative when seasonal swings grow with the level. |
| `changepoint_prior_scale` | numeric | `0.05` | Controls trend flexibility. Larger values allow more changepoints (may overfit). |
| `...` | — | — | Any argument accepted by `prophet::prophet()`. |

```{r prophet}
prophet_fct <- milt_model("prophet",
                           yearly_seasonality      = TRUE,
                           seasonality_mode        = "multiplicative",
                           changepoint_prior_scale = 0.05) |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)
```

---

# Machine learning models

All ML models convert the time series into a **supervised learning** problem by
building lag features from the training window. The target is the next-step
value; the features are lagged values at positions defined by the `lags`
argument.

## `xgboost` — XGBoost

Gradient-boosted decision trees.  Fast, accurate, handles non-linearities and
interactions automatically.

**Requires:** `xgboost`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `lags` | integer vector | `1:12` | Lag indices used as features (e.g., `1:12` uses lags 1 through 12). |
| `nrounds` | integer | `100L` | Number of boosting rounds. Increase for more complex patterns. |
| `max_depth` | integer | `6L` | Maximum tree depth. Deeper trees capture more interactions but may overfit. |
| `eta` | numeric | `0.1` | Learning rate / shrinkage. Lower values need more `nrounds`. |
| `nthread` | integer | `1L` | Number of CPU threads for training. |
| `...` | — | — | Any argument accepted by `xgboost::xgb.train()`. |

```{r xgboost}
xgb_fct <- milt_model("xgboost",
                        lags      = 1:24,
                        nrounds   = 200L,
                        max_depth = 6L,
                        eta       = 0.05) |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

**Feature importance** is available via `milt_explain()`:

```{r xgb-explain}
fitted_xgb <- milt_model("xgboost", lags = 1:12) |> milt_fit(air)
ex <- milt_explain(fitted_xgb, air)
plot(ex)
```

---

## `lightgbm` — LightGBM

Microsoft's gradient-boosting framework.  Faster training and lower memory use
than XGBoost on large datasets, especially with many features.

**Requires:** `lightgbm`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `lags` | integer vector | `1:12` | Lag indices used as features. |
| `num_iterations` | integer | `100L` | Number of boosting rounds. |
| `learning_rate` | numeric | `0.1` | Learning rate (shrinkage). |
| `num_leaves` | integer | `31L` | Maximum number of leaves per tree. Increase for more complex models. |
| `num_threads` | integer | `1L` | Number of CPU threads. |
| `...` | — | — | Any argument accepted by `lightgbm::lgb.train()`. |

```{r lightgbm}
lgb_fct <- milt_model("lightgbm",
                        lags           = 1:12,
                        num_iterations = 200L,
                        learning_rate  = 0.05,
                        num_leaves     = 63L) |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

---

## `random_forest` — Random Forest

Ensemble of decision trees via `ranger`.  More stable than single trees;
naturally handles interactions and outliers.

**Requires:** `ranger`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `lags` | integer vector | `1:12` | Lag indices used as features. |
| `num.trees` | integer | `500L` | Number of trees. More trees = more stable but slower. |
| `mtry` | integer or NULL | `NULL` | Features tried at each split. `NULL` uses `floor(sqrt(p))`. |
| `min.node.size` | integer | `5L` | Minimum terminal node size. Larger values regularise. |
| `num.threads` | integer | `1L` | Number of CPU threads. |
| `importance` | character | `"impurity"` | Variable importance: `"impurity"` (Gini) or `"permutation"`. |
| `...` | — | — | Any argument accepted by `ranger::ranger()`. |

```{r random-forest}
rf_fct <- milt_model("random_forest",
                      lags         = 1:12,
                      num.trees    = 300L,
                      min.node.size = 3L) |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

---

## `elastic_net` — Elastic-Net Regression

Linear model with L1 + L2 regularisation via `glmnet`.  Performs automatic
lag selection when `lambda = NULL` (5-fold CV).  Interpretable and fast.

**Requires:** `glmnet`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `lags` | integer vector | `1:12` | Lag indices used as features. |
| `alpha` | numeric in [0, 1] | `0.5` | Mixing parameter: `0` = ridge, `1` = lasso, `0.5` = elastic net. |
| `lambda` | numeric or NULL | `NULL` | Regularisation strength. `NULL` selects by 5-fold cross-validation. |
| `...` | — | — | Any argument accepted by `glmnet::glmnet()`. |

```{r elastic-net}
# Lasso (alpha = 1) — maximum sparsity
lasso_fct <- milt_model("elastic_net",
                          lags  = 1:24,
                          alpha = 1.0) |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

---

## `svm` — Support Vector Machine

Regression SVM via `e1071`.  Works well on small-to-medium datasets with
non-linear patterns.

**Requires:** `e1071`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `lags` | integer vector | `1:12` | Lag indices used as features. |
| `kernel` | character | `"radial"` | Kernel: `"radial"` (RBF), `"linear"`, `"polynomial"`, `"sigmoid"`. |
| `cost` | numeric | `1` | Regularisation cost `C`. Larger values fit training data more tightly. |
| `epsilon` | numeric | `0.1` | Epsilon-insensitive loss tube width. |
| `gamma` | numeric or NULL | `NULL` | Kernel coefficient. `NULL` uses `1 / n_features`. |

```{r svm}
svm_fct <- milt_model("svm",
                        lags    = 1:12,
                        kernel  = "radial",
                        cost    = 10,
                        epsilon = 0.05) |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

---

## `knn` — K-Nearest Neighbours

Non-parametric lag-based forecasting.  Finds the `k` most similar historical
windows and averages their next steps.

**No package dependency.**

| Parameter | Type | Default | Description |
|---|---|---|---|
| `k` | integer | `5L` | Number of nearest neighbours. |
| `lags` | integer vector | `1:12` | Lag indices used as features. |
| `weights` | character | `"uniform"` | Weighting: `"uniform"` (equal weight) or `"distance"` (inverse-distance). |

```{r knn}
knn_fct <- milt_model("knn",
                        k       = 7L,
                        lags    = 1:12,
                        weights = "distance") |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

---

# Deep learning models

All deep learning backends are built with [torch for R](https://torch.mlverse.org/).
They require `torch` to be installed **and** the Lantern C++ backend downloaded:

```{r torch-setup}
milt_install_backends("deep_learning")
torch::install_torch()   # run once after installing torch
```

All deep learning models share these training parameters:

| Parameter | Description |
|---|---|
| `n_epochs` | Maximum training epochs |
| `lr` | Adam learning rate |
| `patience` | Early-stopping patience (no-improvement epochs) |
| `val_split` | Fraction of windows held out for validation |

---

## `nbeats` — N-BEATS

Neural Basis Expansion Analysis (Oreshkin et al. 2019).  Pure multi-layer
perceptron architecture with stacks of fully-connected blocks.  Each block
produces a *backcast* (removing information from the input) and a *forecast*;
block outputs are summed.  Achieves state-of-the-art results on M4 competition
without domain knowledge.

**Requires:** `torch`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `input_chunk_length` | integer | `24L` | Lookback window length fed to the network. Should cover at least 2 seasonal cycles. |
| `output_chunk_length` | integer | `12L` | Number of steps produced per forward pass. Recursive multi-step if `horizon > output_chunk_length`. |
| `n_stacks` | integer | `2L` | Number of stacks. |
| `n_blocks` | integer | `3L` | Blocks per stack. |
| `hidden_size` | integer | `64L` | FC layer width. |
| `n_layers` | integer | `4L` | FC layer depth per block. |
| `n_epochs` | integer | `100L` | Maximum training epochs. |
| `lr` | numeric | `1e-3` | Adam learning rate. |
| `patience` | integer | `10L` | Early-stopping patience. |
| `val_split` | numeric | `0.1` | Validation fraction. |

```{r nbeats}
nbeats_fct <- milt_model("nbeats",
                           input_chunk_length  = 24L,
                           output_chunk_length = 12L,
                           n_stacks            = 2L,
                           hidden_size         = 128L,
                           n_epochs            = 200L,
                           patience            = 15L) |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)
```

---

## `nhits` — N-HiTS

Neural Hierarchical Interpolation for Time Series (Challu et al. 2022).
Extends N-BEATS with multi-rate sampling (pooling) and hierarchical
interpolation, reducing computation while improving long-horizon accuracy.

**Requires:** `torch`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `input_chunk_length` | integer | `24L` | Lookback window length. |
| `output_chunk_length` | integer | `12L` | Steps predicted per forward pass. |
| `n_stacks` | integer | `3L` | Number of hierarchical stacks (each operates at a different temporal resolution). |
| `n_blocks` | integer | `1L` | Blocks per stack. |
| `hidden_size` | integer | `64L` | FC layer width. |
| `n_layers` | integer | `4L` | FC layer depth per block. |
| `n_epochs` | integer | `100L` | Maximum training epochs. |
| `lr` | numeric | `1e-3` | Adam learning rate. |
| `patience` | integer | `10L` | Early-stopping patience. |
| `val_split` | numeric | `0.1` | Validation fraction. |

```{r nhits}
nhits_fct <- milt_model("nhits",
                          input_chunk_length  = 36L,
                          output_chunk_length = 12L,
                          n_stacks            = 3L) |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)
```

---

## `tft` — Temporal Fusion Transformer

Combines LSTMs and multi-head attention to capture both short-term patterns
(via LSTM encoder) and long-term dependencies (via self-attention).  Designed
for multi-horizon probabilistic forecasting.

**Requires:** `torch`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `input_chunk_length` | integer | `24L` | Lookback window fed to the LSTM encoder. |
| `output_chunk_length` | integer | `12L` | Steps predicted per forward pass. |
| `hidden_size` | integer | `64L` | LSTM hidden state and attention dimension. |
| `n_heads` | integer | `4L` | Number of attention heads. |
| `n_layers` | integer | `2L` | LSTM encoder depth. |
| `dropout` | numeric | `0.1` | Dropout rate. |
| `n_epochs` | integer | `100L` | Maximum training epochs. |
| `lr` | numeric | `1e-3` | Adam learning rate. |
| `patience` | integer | `10L` | Early-stopping patience. |
| `val_split` | numeric | `0.1` | Validation fraction. |

```{r tft}
tft_fct <- milt_model("tft",
                        input_chunk_length  = 36L,
                        output_chunk_length = 12L,
                        hidden_size         = 128L,
                        n_heads             = 4L,
                        dropout             = 0.2,
                        n_epochs            = 150L) |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)
```

---

## `tcn` — Temporal Convolutional Network

Dilated causal convolutions allow the receptive field to grow exponentially
with depth, capturing long-range dependencies without recurrence.

**Requires:** `torch`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `input_chunk_length` | integer | `24L` | Lookback window length. |
| `output_chunk_length` | integer | `12L` | Steps predicted per forward pass. |
| `n_filters` | integer | `32L` | Number of convolutional channels per layer. |
| `kernel_size` | integer | `3L` | Convolution kernel size. |
| `n_layers` | integer | `4L` | Number of temporal residual blocks. |
| `dropout` | numeric | `0.2` | Dropout rate in residual blocks. |
| `n_epochs` | integer | `100L` | Maximum training epochs. |
| `lr` | numeric | `1e-3` | Adam learning rate. |
| `patience` | integer | `10L` | Early-stopping patience. |
| `val_split` | numeric | `0.1` | Validation fraction. |

```{r tcn}
tcn_fct <- milt_model("tcn",
                        input_chunk_length  = 24L,
                        output_chunk_length = 12L,
                        n_filters           = 64L,
                        n_layers            = 6L,
                        dropout             = 0.1) |>
  milt_fit(air) |>
  milt_forecast(horizon = 12)
```

---

## `patch_tst` — PatchTST

Applies a Vision Transformer concept to time series: the input window is split
into non-overlapping **patches**, each patch is linearly embedded into a token,
learnable positional encodings are added, and a standard Transformer encoder
processes all patch tokens jointly.  Excels at capturing long-range
dependencies efficiently.

**Requires:** `torch`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `input_chunk_length` | integer | `24L` | Lookback window length. |
| `output_chunk_length` | integer | `12L` | Steps predicted per forward pass. |
| `patch_len` | integer | `8L` | Length of each non-overlapping patch. Must be ≤ `input_chunk_length`. |
| `d_model` | integer | `64L` | Patch embedding / Transformer model dimension. |
| `n_heads` | integer | `4L` | Number of attention heads. `d_model` must be divisible by `n_heads`. |
| `n_layers` | integer | `2L` | Transformer encoder depth. |
| `dropout` | numeric | `0.1` | Dropout rate. |
| `n_epochs` | integer | `100L` | Maximum training epochs. |
| `lr` | numeric | `1e-3` | Adam learning rate. |
| `patience` | integer | `10L` | Early-stopping patience. |
| `val_split` | numeric | `0.1` | Validation fraction. |

```{r patch-tst}
ptst_fct <- milt_model("patch_tst",
                         input_chunk_length  = 48L,
                         output_chunk_length = 12L,
                         patch_len           = 8L,
                         d_model             = 128L,
                         n_heads             = 4L,
                         n_layers            = 3L) |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)
```

---

## `deepar` — DeepAR

Probabilistic LSTM-based model trained as an autoregressive likelihood model.
Returns Gaussian parameters (mean and standard deviation) at each step, which
are used to construct prediction intervals analytically.

**Requires:** `torch`

| Parameter | Type | Default | Description |
|---|---|---|---|
| `input_chunk_length` | integer | `24L` | Lookback window fed to the LSTM. |
| `output_chunk_length` | integer | `12L` | Steps predicted per forward pass. |
| `hidden_size` | integer | `64L` | LSTM hidden state dimension. |
| `n_layers` | integer | `2L` | Number of LSTM layers. |
| `n_epochs` | integer | `100L` | Maximum training epochs. |
| `lr` | numeric | `1e-3` | Adam learning rate. |
| `patience` | integer | `10L` | Early-stopping patience. |
| `val_split` | numeric | `0.1` | Validation fraction. |

```{r deepar}
deepar_fct <- milt_model("deepar",
                           input_chunk_length  = 36L,
                           output_chunk_length = 12L,
                           hidden_size         = 128L,
                           n_layers            = 3L,
                           n_epochs            = 200L) |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)
```

---

# Comparing models

Use `milt_compare()` to run a rolling-origin backtest across multiple models
and rank them by accuracy:

```{r compare}
models <- list(
  naive  = milt_model("naive"),
  arima  = milt_model("auto_arima"),
  ets    = milt_model("ets"),
  xgb    = milt_model("xgboost", lags = 1:12)
)

cmp <- milt_compare(models, air, horizon = 12, rank_metric = "RMSE")
print(cmp)
plot(cmp)
```

---

# Ensembling

Combine fitted models into a weighted or mean ensemble:

```{r ensemble}
m1 <- milt_model("auto_arima") |> milt_fit(air)
m2 <- milt_model("ets")        |> milt_fit(air)
m3 <- milt_model("xgboost", lags = 1:12) |> milt_fit(air)

ens <- milt_ensemble(list(arima = m1, ets = m2, xgb = m3), method = "mean")
ens_fct <- milt_forecast(ens, horizon = 12)
plot(ens_fct)
```
