---
title: "Anomaly Detection with milt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anomaly Detection with milt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(milt)
```

## Overview

milt's anomaly detection API follows a two-step pattern:

```r
detector <- milt_detector("<method>", ...)
anomalies <- milt_detect(detector, series)
```

Results are returned as a `MiltAnomalies` object with `print()`, `plot()`,
and `as_tibble()` methods.

---

## 1. Create a series with injected spikes

```{r spike-series}
set.seed(42)
tbl <- tibble::tibble(
  date  = seq(as.Date("2020-01-01"), by = "month", length.out = 60),
  value = c(rnorm(20, 10, 1), 50, rnorm(19, 10, 1),   # spike at 21
             rnorm(10, 10, 1), -30, rnorm(8, 10, 1))   # dip at 51
)
s <- milt_series(tbl, time_col = "date", value_cols = "value")
plot(s)
```

---

## 2. IQR detector (fast, no assumptions)

```{r iqr}
d   <- milt_detector("iqr", k = 1.5)
anm <- milt_detect(d, s)
print(anm)
plot(anm)
```

---

## 3. GESD test (statistical, iterative)

```{r gesd}
d_gesd <- milt_detector("gesd", max_anoms = 5, alpha = 0.05)
anm_gesd <- milt_detect(d_gesd, s)
plot(anm_gesd)
```

---

## 4. STL decomposition–based

STL separates trend and seasonality; anomalies are in the remainder:

```{r stl}
air <- milt_series(AirPassengers)
d_stl  <- milt_detector("stl", threshold = 3)
anm_stl <- milt_detect(d_stl, air)
print(anm_stl)
plot(anm_stl)
```

---

## 5. Isolation Forest

Requires the `isotree` package:

```{r iforest, eval = requireNamespace("isotree", quietly = TRUE)}
d_if  <- milt_detector("iforest", n_trees = 50, threshold = 0.6)
anm_if <- milt_detect(d_if, s)
plot(anm_if)
```

---

## 6. Local Outlier Factor

Requires the `dbscan` package:

```{r lof, eval = requireNamespace("dbscan", quietly = TRUE)}
d_lof  <- milt_detector("lof", k = 5, threshold = 1.5)
anm_lof <- milt_detect(d_lof, s)
plot(anm_lof)
```

---

## 7. Ensemble detection

Combine multiple detectors with majority vote or mean score:

```{r ensemble}
d1  <- milt_detector("iqr",   k = 1.5)
d2  <- milt_detector("grubbs", alpha = 0.05)
d3  <- milt_detector("gesd",  max_anoms = 5)

ens <- milt_detector("ensemble",
                      detectors = list(d1, d2, d3),
                      method    = "majority")
anm_ens <- milt_detect(ens, s)
plot(anm_ens)
```

---

## 8. Changepoint detection

```{r changepoints, eval = requireNamespace("changepoint", quietly = TRUE)}
cp <- milt_changepoints(air, method = "pelt", stat = "mean")
print(cp)
plot(cp)
```

---

## 9. Tidy output

```{r tidy}
tbl_out <- tibble::as_tibble(milt_detect(milt_detector("iqr"), s))
head(tbl_out)
```
