---
title: "EDA and Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDA and Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

milt provides two complementary tools for understanding a time series before
modelling:

- **`milt_eda()`** — a *statistical* deep-dive: descriptive statistics,
  stationarity tests, seasonality strength, ACF/PACF, decomposition, and
  distribution plots all in one call.
- **`milt_diagnose()`** — a *diagnostic* summary: stationarity, seasonality,
  trend, gap detection, outlier count, and actionable modelling recommendations.

Both functions accept a `MiltSeries` object and are fully pipe-friendly.

---

## 1. Quick Start

```{r}
library(milt)

ap <- milt_series(AirPassengers)

eda  <- milt_eda(ap)
diag <- milt_diagnose(ap)

print(eda)
print(diag)
```

---

## 2. Exploratory Data Analysis — `milt_eda()`

```r
milt_eda(series, plots = TRUE, quiet = FALSE)
```

| Argument | Type | Default | Description |
|---|---|---|---|
| `series` | MiltSeries | — | The time series to analyse. |
| `plots` | logical | `TRUE` | Whether to render ACF/PACF, decomposition, and distribution plots. Set to `FALSE` for non-interactive or batch use. |
| `quiet` | logical | `FALSE` | Suppress cli progress messages during analysis. |

**Returns:** a `MiltEDA` object. Use `print()`, `summary()`, or
`tibble::as_tibble()` to inspect results.

### 2.1 Descriptive Statistics

The EDA report starts with a table of basic statistics:

```{r}
eda <- milt_eda(ap)
print(eda)
```

Reported statistics include:

| Statistic | Description |
|---|---|
| `n` | Total number of observations. |
| `mean` | Arithmetic mean of values. |
| `sd` | Standard deviation. |
| `min` / `max` | Range of the series. |
| `q25` / `q75` | 25th and 75th percentiles (IQR). |
| `missing` | Count and percentage of `NA` values. |
| `zeros` | Count of exact zero values. |

Extract as a tibble for downstream use:

```{r}
tibble::as_tibble(eda)
```

### 2.2 Stationarity Tests

milt automatically runs two complementary stationarity tests:

| Test | H₀ | Pass condition | Method |
|---|---|---|---|
| **PP (Phillips-Perron)** | Unit root present (non-stationary) | p-value < 0.05 rejects H₀ → stationary | `stats::PP.test()` |
| **KPSS (Kwiatkowski-Phillips-Schmidt-Shin)** | Level stationary | p-value ≥ 0.05 fails to reject H₀ → stationary | Base-R implementation (Newey-West LRV) |

Both tests are always computed using base-R implementations (no extra
packages required).

**Interpreting the combined result:**

| PP p-value | KPSS p-value | Conclusion |
|---|---|---|
| < 0.05 | ≥ 0.05 | Stationary — no differencing needed. |
| ≥ 0.05 | < 0.05 | Non-stationary — consider first-differencing. |
| < 0.05 | < 0.05 | Ambiguous — inspect ACF/PACF manually. |
| ≥ 0.05 | ≥ 0.05 | Ambiguous — likely a borderline case. |

Access raw test results:

```{r}
stat <- eda$stationarity()

stat$pp_pvalue    # Phillips-Perron p-value
stat$kpss_pvalue  # KPSS p-value
stat$stationary   # logical: TRUE if both tests agree series is stationary
```

### 2.3 Seasonality Detection

Seasonal strength is computed via STL decomposition
(`stats::stl(s.window = "periodic", robust = TRUE)`):

$$\text{strength} = \max\!\left(0,\ 1 - \frac{\text{Var(remainder)}}{\text{Var(seasonal)} + \text{Var(remainder)}}\right)$$

Values range from 0 (no seasonality) to 1 (perfectly seasonal). A strength
above **0.40** is generally considered meaningful for modelling purposes.

```{r}
seas <- eda$seasonality()

seas$period    # dominant seasonal period (e.g. 12 for monthly data)
seas$strength  # strength in [0, 1]
seas$seasonal  # logical: TRUE if strength > 0.40
```

### 2.4 Visual Outputs

When `plots = TRUE`, `milt_eda()` renders the following plots:

**ACF / PACF** — generated by `milt_plot_acf()`:

```{r}
milt_plot_acf(ap, lags = 36)
```

Useful for identifying:
- Slow ACF decay → non-stationary or trend present.
- Spikes at lag *m* → seasonality with period *m*.
- Significant PACF at lag 1–2 only → AR process.

**Decomposition** — generated by `milt_plot_decomp()`:

```{r}
milt_plot_decomp(ap)
```

Decomposes the series into trend, seasonal, and remainder components using
STL. Helpful for understanding the dominant patterns before choosing a model.

**Distribution** — a histogram + density overlay of the value distribution.
Helps identify skewness, heavy tails, or bimodality that may warrant
transformations (e.g. log, Box-Cox).

---

## 3. Series Diagnostics — `milt_diagnose()`

```r
milt_diagnose(series)
```

| Argument | Type | Description |
|---|---|---|
| `series` | MiltSeries | The time series to diagnose. |

**Returns:** a `MiltDiagnosis` object. Use `print()` or `summary()`.

### 3.1 What `milt_diagnose()` checks

`milt_diagnose()` runs six checks and synthesises modelling recommendations:

#### Stationarity
Uses the coefficient-of-variation ratio: if the rolling-window CV is stable
across the series, it is classified as stationary. The result is a logical
`$stationary` flag plus a short recommendation.

#### Seasonality
Uses STL-based strength (identical to `milt_eda()`):

```{r}
diag <- milt_diagnose(ap)

diag$.__enclos_env__$private$.seasonality$seasonal   # TRUE/FALSE
diag$.__enclos_env__$private$.seasonality$strength   # numeric
diag$.__enclos_env__$private$.seasonality$period     # integer
```

#### Trend
A simple linear regression of values on time. The slope and its p-value
determine whether a trend is considered present (`p < 0.05`):

```{r}
diag$.__enclos_env__$private$.trend$has_trend  # logical
diag$.__enclos_env__$private$.trend$slope      # regression slope
diag$.__enclos_env__$private$.trend$p_value    # linear trend p-value
```

#### Gap Detection
Checks whether consecutive time steps match the expected frequency. Reports
a tibble of gap locations:

```{r}
gaps <- diag$.__enclos_env__$private$.gaps
# A tibble: time_before, time_after, n_missing
```

#### Outlier Detection
Uses a simple IQR rule: values outside
$[Q1 - 3 \times \text{IQR},\, Q3 + 3 \times \text{IQR}]$ are flagged.

```{r}
out <- diag$.__enclos_env__$private$.outliers
out$n_outliers  # integer count
out$indices     # integer vector of row indices
```

#### Recommendations
Based on all checks, `milt_diagnose()` returns a character vector of
actionable suggestions:

```{r}
diag$.__enclos_env__$private$.recommendations
```

Example output for `AirPassengers`:

```
[1] "Series appears non-stationary — consider differencing or a model with trend/drift."
[2] "Strong seasonality detected (period = 12) — use a seasonal model (snaive, ETS, SARIMA)."
[3] "Trend present — include trend component or apply detrending."
```

### 3.2 Full Diagnostic Report

```{r}
diag <- milt_diagnose(ap)
print(diag)

# MiltDiagnosis
# Series    : 144 obs @ monthly
# Range     : 1949-01-01 — 1960-12-01
#
# Stationarity : non-stationary (CV ratio = 0.412)
# Seasonality  : seasonal (strength = 0.92, period = 12)
# Trend        : trend present (slope = 2.667, p = 3.1e-52)
# Gaps         : 0
# Outliers     : 0
#
# Recommendations:
#  • Consider differencing or a model with trend/drift.
#  • Use a seasonal model (snaive, ETS, SARIMA, TBATS, Prophet).
```

---

## 4. ACF and PACF Plots — `milt_plot_acf()`

```r
milt_plot_acf(series, lags = NULL, type = c("both", "acf", "pacf"))
```

| Argument | Type | Default | Description |
|---|---|---|---|
| `series` | MiltSeries | — | Input series. |
| `lags` | positive integer or NULL | `NULL` | Number of lags to plot. Defaults to `min(floor(n / 4), 50)`. |
| `type` | character | `"both"` | Which plot(s) to render: `"acf"`, `"pacf"`, or `"both"` (side-by-side). |

**Reading the ACF:**

- **Slow geometric decay** (all lags positive and declining slowly) →
  strong trend; series needs differencing.
- **Sine-wave pattern** → deterministic seasonality.
- **Single significant spike at lag *k*, then cuts off** → MA(*k*) process.
- **All lags inside confidence bands** → white noise (no autocorrelation).

**Reading the PACF:**

- **Single significant spike at lag 1** → AR(1) process.
- **Significant spikes at lags 1–*p*, then cuts off** → AR(*p*) process.
- **Spike at seasonal lag *m*** → seasonal AR term.

```{r}
# Side-by-side ACF and PACF
milt_plot_acf(ap, lags = 36, type = "both")

# ACF only with custom lag count
milt_plot_acf(ap, lags = 24, type = "acf")
```

---

## 5. Decomposition Plot — `milt_plot_decomp()`

```r
milt_plot_decomp(series, method = c("stl", "classical"), robust = TRUE)
```

| Argument | Type | Default | Description |
|---|---|---|---|
| `series` | MiltSeries | — | Input series. Must have a known seasonal period > 1. |
| `method` | character | `"stl"` | Decomposition method: `"stl"` (STL via `stats::stl()`) or `"classical"` (additive/multiplicative via `stats::decompose()`). |
| `robust` | logical | `TRUE` | Use robust fitting in STL (less sensitive to outliers). Only applies when `method = "stl"`. |

**Returns:** a ggplot2 object with four panels: observed, trend, seasonal,
and remainder.

```{r}
milt_plot_decomp(ap)

# Inspect the STL remainder for anomalies
milt_plot_decomp(ap, method = "stl", robust = TRUE)
```

---

## 6. Complete Pre-Modelling Workflow

The recommended workflow before fitting any model:

```{r}
library(milt)

# 1. Load data
ap <- milt_series(AirPassengers)

# 2. Diagnose structure
diag <- milt_diagnose(ap)
print(diag)

# 3. Deep-dive EDA
eda <- milt_eda(ap)
print(eda)

# 4. Visualise autocorrelation structure
milt_plot_acf(ap, lags = 36)

# 5. Decompose
milt_plot_decomp(ap)

# 6. Choose model based on findings
# Strong seasonality + trend → ETS, SARIMA, or Prophet
fc <- milt_model("auto_arima") |>
  milt_fit(ap) |>
  milt_forecast(h = 24)

plot(fc)
```

### Decision tree based on diagnostic output

| Finding | Recommended action |
|---|---|
| Non-stationary (trend) | Use ARIMA with differencing, ETS(A,A,*), or Prophet |
| Seasonal + trend | auto_arima, ETS(A,A,A), TBATS, Prophet |
| Seasonal only | snaive baseline, ETS(A,N,A), SARIMA |
| Stationary, no seasonality | ARIMA(p,0,q), ETS(A,N,N) |
| Irregular gaps | `milt_fill_gaps()` before fitting |
| Many outliers | Robust fitting; consider `robust = TRUE` in decomp |
| Short series (< 2 seasons) | Naïve or drift; avoid complex models |

---

## See Also

- `vignette("milt-series-api")` — creating and manipulating MiltSeries
- `vignette("milt-forecasting")` — fitting and forecasting models
- `vignette("milt-model-reference")` — model parameter reference
- `vignette("milt-metrics-guide")` — accuracy metrics reference
