---
title: "Forecasting with milt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting with milt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = requireNamespace("forecast", quietly = TRUE)
)
```

```{r setup}
library(milt)
```

## Overview

This vignette walks through milt's full forecasting model zoo — from classical
statistical models to machine learning and ensembles — using a single
consistent API.

```r
milt_model("<name>") |> milt_fit(series) |> milt_forecast(horizon = 12)
```

---

## 1. The model zoo

List every registered model:

```{r list-models}
list_milt_models()
```

---

## 2. Classical models

```{r classical}
air <- milt_series(AirPassengers)

# Auto-ARIMA
fct_arima <- milt_model("auto_arima") |> milt_fit(air) |> milt_forecast(24)
plot(fct_arima)

# ETS
fct_ets <- milt_model("ets") |> milt_fit(air) |> milt_forecast(24)
plot(fct_ets)
```

---

## 3. Train/test evaluation

```{r train-test}
spl <- milt_split(air, ratio = 0.8)
h   <- spl$test$n_timesteps()

fct_cv <- milt_model("ets") |>
  milt_fit(spl$train) |>
  milt_forecast(h)

milt_accuracy(spl$test$values(), fct_cv$as_tibble()$.mean)
```

---

## 4. ML models with lag features

XGBoost, LightGBM, random forest, and elastic net all use automatic lag
features.  You can customise the lags:

```{r ml-models, eval = requireNamespace("xgboost", quietly = TRUE)}
fct_xgb <- milt_model("xgboost", lags = 1:12) |>
  milt_fit(air) |>
  milt_forecast(12)
plot(fct_xgb)
```

---

## 5. Model comparison

`milt_compare()` runs a rolling-origin backtest on each model and ranks them:

```{r compare, eval = requireNamespace("xgboost", quietly = TRUE)}
models <- list(
  arima  = milt_model("auto_arima"),
  ets    = milt_model("ets"),
  xgb    = milt_model("xgboost", lags = 1:12)
)
cmp <- milt_compare(models, air, horizon = 12)
print(cmp)
plot(cmp)
```

---

## 6. Ensemble models

Combine forecasts with a mean ensemble:

```{r ensemble}
m1 <- milt_model("auto_arima") |> milt_fit(air)
m2 <- milt_model("ets")        |> milt_fit(air)
m3 <- milt_model("naive")      |> milt_fit(air)

ens <- milt_ensemble(list(m1, m2, m3), method = "mean")
fct_ens <- milt_forecast(ens, 12)
plot(fct_ens)
```

---

## 7. Probabilistic forecasts

Every model returns 80% and 95% prediction intervals.  Access them via
`as_tibble()`:

```{r prob-forecast}
tbl <- fct_arima$as_tibble()
head(tbl)
```

For sample-path–based probabilistic forecasts, use `num_samples`:

```{r samples, eval = FALSE}
fct_samples <- milt_model("naive") |>
  milt_fit(air) |>
  milt_forecast(12, num_samples = 200)
```

---

## 8. Backtesting

`milt_backtest()` implements expanding- and sliding-window CV:

```{r backtest}
bt <- milt_backtest(
  milt_model("ets"),
  series  = air,
  horizon = 12,
  stride  = 12,
  method  = "expanding"
)
print(bt)
plot(bt)
```
