---
title: "Building Pipelines with milt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Pipelines with milt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(milt)
```

## Overview

Feature engineering in milt is done with `milt_step_*()` functions that each
accept and return a `MiltSeries`.  They compose naturally with the native
pipe `|>`.

---

## 1. Lag features

```{r lags}
air    <- milt_series(AirPassengers)
air_lg <- milt_step_lag(air, lags = 1:12)
print(air_lg)
```

---

## 2. Rolling statistics

```{r rolling}
air_roll <- milt_step_rolling(air, windows = c(3, 6, 12), fns = c("mean", "sd"))
print(air_roll)
```

---

## 3. Fourier terms

Fourier terms capture complex seasonality:

```{r fourier}
air_f <- milt_step_fourier(air, period = 12, K = 3)
print(air_f)
```

---

## 4. Calendar features

```{r calendar}
tbl <- tibble::tibble(
  date  = seq(as.Date("2020-01-01"), by = "day", length.out = 365),
  value = rnorm(365, 100, 10)
)
s_daily <- milt_series(tbl, time_col = "date", value_cols = "value")

s_cal <- milt_step_calendar(s_daily)
print(s_cal)
```

---

## 5. Scaling

`milt_step_scale()` normalises values and returns an invertible step object:

```{r scaling}
out       <- milt_step_scale(air, method = "zscore")
air_z     <- out$series    # scaled MiltSeries
scaler    <- out$step      # MiltScaleStep — keeps centering/scaling params

# Verify: mean ≈ 0, sd ≈ 1
cat("Mean:", round(mean(air_z$values()), 4), "\n")
cat("SD:  ", round(sd(air_z$values()),   4), "\n")

# Invert
air_back  <- milt_step_unscale(scaler, air_z)
cat("Original max:", max(air$values()), "\n")
cat("Restored max:", max(air_back$values()), "\n")
```

---

## 6. Chaining steps

Steps compose with `|>`:

```{r chain}
air_features <- air |>
  milt_step_lag(lags = 1:6)   |>
  milt_step_rolling(windows = 3, fns = "mean") |>
  milt_step_fourier(period = 12, K = 2)

print(air_features)
```

---

## 7. Using features with an ML model

```{r ml-pipeline, eval = requireNamespace("xgboost", quietly = TRUE)}
# Scale → fit → forecast
scaled_out <- milt_step_scale(air, method = "zscore")
air_scaled <- scaled_out$series
scaler     <- scaled_out$step

m   <- milt_model("xgboost", lags = 1:12) |> milt_fit(air_scaled)
fct <- milt_forecast(m, 12)

# Point forecasts back in original units
fct_vals_orig <- scaler$inverse_transform_vector(fct$as_tibble()$.mean)
cat("Forecast (original units):\n")
print(round(fct_vals_orig, 1))
```

---

## 8. Model explainability

After fitting an ML model, inspect feature importance:

```{r explain, eval = requireNamespace("xgboost", quietly = TRUE)}
m   <- milt_model("xgboost", lags = 1:12) |> milt_fit(air)
exp <- milt_explain(m)
print(exp)
plot(exp)
```
