---
title: "Deploying milt Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deploying milt Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = FALSE   # deployment examples require interactive sessions
)
```

```{r setup}
library(milt)
```

## Overview

Once a model is trained it can be:
1. **Saved to disk** with `milt_save()` and restored with `milt_load()`.
2. **Served as a REST API** with `milt_serve()` (Plumber).
3. **Explored in a dashboard** with `milt_dashboard()` (Shiny).
4. **Included in a report** with `milt_report()` (R Markdown / Quarto).

---

## 1. Save and load

```{r save-load}
air <- milt_series(AirPassengers)
m   <- milt_model("ets") |> milt_fit(air)

# Save
path <- milt_save(m, "my_ets_model")   # writes my_ets_model.milt
cat("Saved to:", path, "\n")

# Load later
m2 <- milt_load("my_ets_model.milt")
fct <- milt_forecast(m2, 12)
plot(fct)
```

Any milt object can be saved: `MiltSeries`, `MiltForecast`, `MiltAnomalies`,
detectors, etc.

---

## 2. REST API with Plumber

`milt_serve()` wraps the model in a Plumber API that exposes:

| Endpoint        | Method | Description                        |
|-----------------|--------|------------------------------------|
| `/health`       | GET    | Version + model name               |
| `/series_info`  | GET    | Training series metadata           |
| `/forecast`     | POST   | Accepts `{"horizon": <int>}`, returns JSON forecast |

```{r serve}
# Launch (blocking — runs until interrupted)
milt_serve(m, host = "127.0.0.1", port = 8000)

# Non-blocking: get the router object without starting
router <- milt_serve(m, launch = FALSE)
```

Test with curl:

```bash
curl -s http://127.0.0.1:8000/health
curl -s -X POST http://127.0.0.1:8000/forecast \
     -H "Content-Type: application/json" \
     -d '{"horizon": 12}'
```

---

## 3. Shiny monitoring dashboard

`milt_dashboard()` launches an interactive four-tab app:

* **Series** — interactive line chart of training data.
* **Forecast** — adjustable horizon slider with PI fan chart.
* **Diagnostics** — residuals, ACF plot, and histogram.
* **Anomalies** — IQR anomaly overlay with adjustable multiplier.

```{r dashboard}
milt_dashboard(m)
```

---

## 4. Automated reports

`milt_report()` renders a self-contained HTML or PDF analysis report:

```{r report}
# HTML (default)
path <- milt_report(
  series  = air,
  models  = list(ets = m),
  horizon = 24
)
cat("Report at:", path, "\n")

# PDF
milt_report(air, output_format = "pdf", output_file = "air_report.pdf")
```

The report includes:
- Series overview and plot
- Diagnostic tests (ADF, KPSS, seasonality)
- EDA statistics
- Forecast plots for each supplied model
- Anomaly detection overview

---

## 5. Model versioning

Save checkpoints at each refit:

```{r versioning}
# After initial training
milt_save(m, "model_v1")

# After refit on new data
new_data <- milt_series(AirPassengers)
m_v2     <- milt_refit(m, new_data)
milt_save(m_v2, "model_v2")

# Load the best version
best <- milt_load("model_v1.milt")
```
