---
title: "Working with Multiple Series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Multiple Series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(milt)
```

## Overview

milt handles multi-series data natively.  A grouped `MiltSeries` stores
multiple series in a single object, and `milt_local_model()` fits one model
per series in parallel.

---

## 1. Creating a multi-series MiltSeries

```{r multi-create}
# Build a synthetic dataset with 3 stores
set.seed(42)
n   <- 48L
tbl <- tibble::tibble(
  date     = rep(seq(as.Date("2020-01-01"), by = "month", length.out = n), 3),
  store    = rep(c("A", "B", "C"), each = n),
  sales    = c(
    rnorm(n, 100, 10),
    rnorm(n, 200, 20),
    rnorm(n, 150, 15)
  )
)

ms <- milt_series(tbl,
                   time_col  = "date",
                   value_cols = "sales",
                   group_col = "store")
print(ms)
```

---

## 2. Visualise all series

```{r multi-plot}
plot(ms)
```

---

## 3. Local models — one per series

`milt_local_model()` wraps any model to train independently on each group:

```{r local-model, eval = requireNamespace("forecast", quietly = TRUE)}
local_m <- milt_local_model(milt_model("ets"))
fitted  <- milt_fit(local_m, ms)
fct     <- milt_forecast(fitted, 12)
print(fct)
plot(fct)
```

---

## 4. Split and manipulate individual series

```{r split}
# Extract one group by filtering the underlying tibble
s_A <- milt_series(
  dplyr::filter(ms$as_tibble(), store == "A"),
  time_col   = "date",
  value_cols = "sales"
)

spl <- milt_split(s_A, ratio = 0.8)
cat("Train:", spl$train$n_timesteps(), "Test:", spl$test$n_timesteps(), "\n")
```

---

## 5. Concatenate series

```{r concat}
s1 <- milt_series(AirPassengers)[1:72]
s2 <- milt_series(AirPassengers)[73:144]
s_full <- milt_concat(s1, s2)
cat("Combined length:", s_full$n_timesteps(), "\n")
```

---

## 6. Time series clustering

Group series by shape similarity:

```{r cluster}
# Create a list of separate MiltSeries for clustering
series_list <- lapply(c("A", "B", "C"), function(g) {
  milt_window(ms, group = g)
})

cl <- milt_cluster(series_list, k = 2L, method = "feature_based")
print(cl)
```

---

## 7. Time series classification

Train a classifier on labelled series:

```{r classify, eval = requireNamespace("ranger", quietly = TRUE)}
# Assign synthetic labels
labels <- c("high", "high", "low")   # stores A, B, C

clf <- milt_classifier("feature_based")
milt_classify_fit(clf, series_list, labels)

pred <- milt_classify_predict(clf, series_list)
cat("Predicted labels:", pred$labels, "\n")
```

---

## 8. Hierarchical reconciliation

Ensure aggregate forecasts are coherent with bottom-level sums:

```{r hierarchical, eval = requireNamespace("forecast", quietly = TRUE)}
# Fit models for Total + each store
s_total <- milt_series(
  tibble::tibble(
    date  = seq(as.Date("2020-01-01"), by = "month", length.out = n),
    sales = tbl$sales[tbl$store == "A"] +
            tbl$sales[tbl$store == "B"] +
            tbl$sales[tbl$store == "C"]
  ),
  time_col = "date", value_cols = "sales"
)

fc_total <- milt_model("ets") |> milt_fit(s_total)  |> milt_forecast(6)
fc_A     <- milt_model("ets") |> milt_fit(series_list[[1L]]) |> milt_forecast(6)
fc_B     <- milt_model("ets") |> milt_fit(series_list[[2L]]) |> milt_forecast(6)
fc_C     <- milt_model("ets") |> milt_fit(series_list[[3L]]) |> milt_forecast(6)

# Summing matrix (Total = A + B + C)
S <- matrix(c(1,1,1, 1,0,0, 0,1,0, 0,0,1),
            nrow = 4, ncol = 3, byrow = TRUE,
            dimnames = list(c("Total","A","B","C"), c("A","B","C")))

rc <- milt_reconcile(
  list(Total = fc_total, A = fc_A, B = fc_B, C = fc_C),
  S      = S,
  method = "ols"
)
print(rc)
```
