---
title: "Getting Started with milt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with milt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(milt)
```

## Overview

**milt** (Modern Integrated Library for Timeseries) provides a single,
consistent API for every time series task: forecasting, anomaly detection,
classification, and clustering. Every model follows the same three-step pipe:

```r
milt_model("<name>") |> milt_fit(series) |> milt_forecast(horizon)
```

This vignette walks through the core workflow using the built-in
`AirPassengers` dataset.

---

## 1. Creating a MiltSeries

`milt_series()` accepts virtually any R time series object: a `ts`, a
`data.frame`, a `tibble`, an `xts`, a `tsibble`, or a plain numeric vector.

```{r create-series}
# From a base-R ts object
air <- milt_series(AirPassengers)
print(air)
```

You can also build a series from a tibble and control every option:

```{r create-from-tibble, eval = FALSE}
data(milt_air)
air2 <- milt_series(
  milt_air,
  time_col  = "date",
  value_cols = "value",
  frequency = "monthly"
)
```

Key accessors on a `MiltSeries`:

```{r accessors}
air$n_timesteps()   # number of observations
air$freq()          # detected frequency label
air$start_time()    # first time index
air$end_time()      # last time index
air$values()        # numeric vector of the target variable
```

---

## 2. Diagnosing a series

`milt_diagnose()` runs a quick statistical battery: trend test, stationarity
check, seasonality strength, and outlier count.

```{r diagnose}
dx <- milt_diagnose(air)
print(dx)
```

---

## 3. Fitting and forecasting

Use `milt_model()` to select a model by name, `milt_fit()` to train it, and
`milt_forecast()` to produce point forecasts plus 80% and 95% prediction
intervals.

```{r fit-forecast}
fct <- milt_model("naive") |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)

print(fct)
```

The result is a `MiltForecast` object. Convert it to a tibble for downstream
work:

```{r as-tibble}
head(fct$as_tibble())
```

Plot the forecast with history:

```{r plot-forecast}
plot(fct)
```

### Choosing a model

```{r list-models}
list_milt_models()
```

Swap `"naive"` for any registered key.  Models backed by the `forecast`
package (ETS, ARIMA, Theta, STL) are loaded lazily and only require the
`forecast` package to be installed.

```{r ets-forecast, eval = requireNamespace("forecast", quietly = TRUE)}
fct_ets <- milt_model("ets") |>
  milt_fit(air) |>
  milt_forecast(horizon = 24)

plot(fct_ets)
```

---

## 4. Evaluating accuracy

Split the series into training and test sets, forecast over the test horizon,
and compute accuracy metrics:

```{r accuracy}
spl <- milt_split(air, ratio = 0.8)   # 80 % train, 20 % test

fct_test <- milt_model("naive") |>
  milt_fit(spl$train) |>
  milt_forecast(spl$test$n_timesteps())

acc <- milt_accuracy(
  actual    = spl$test$values(),
  predicted = fct_test$as_tibble()$.mean
)
print(acc)
```

---

## 5. Walk-forward backtesting

`milt_backtest()` performs time series cross-validation without data leakage.
Two strategies are supported:

- **expanding** (default) — the training window grows fold by fold.
- **sliding** — a fixed-size training window advances by `stride` steps.

```{r backtest}
bt <- milt_backtest(
  model          = milt_model("naive"),
  series         = air,
  horizon        = 12,
  initial_window = 108L,
  stride         = 12L,
  metrics        = c("MAE", "RMSE")
)

print(bt)
plot(bt)
```

Retrieve per-fold results as a tibble:

```{r backtest-tibble}
tibble::as_tibble(bt)
```

---

## 6. Handling gaps and missing data

```{r fill-gaps, eval = FALSE}
# Detect and fill gaps with linear interpolation
air_gapped <- air   # hypothetical series with missing values
air_filled <- milt_fill_gaps(air_gapped, method = "linear")
```

Other imputation methods: `"spline"`, `"locf"`, `"nocb"`, `"mean"`, `"zero"`.

---

## 7. Series manipulation

```{r manipulate}
# Subset to 1955-1960
air_55 <- milt_window(air,
                      start = as.Date("1955-01-01"),
                      end   = as.Date("1960-12-01"))
air_55$n_timesteps()

# Head / tail
milt_head(air, 12)$n_timesteps()
milt_tail(air, 12)$n_timesteps()
```

---

## Next steps

- See `vignette("milt-backends")` for a complete model catalogue (coming soon).
- See `vignette("milt-covariates")` for covariate-aware forecasting.
- Visit <https://github.com/ntiGideon/milt> to follow development or open an issue.
